---
title: "NHTSA Data Cleaning"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    fontsize: 0.8em
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 4
    smooth-scroll: TRUE
    embed-resources: TRUE
    fig-align: center
editor: visual
---

# Set-up

## Library

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true  

library(tidyverse)
```

## Import Data

I used Excel to help create functions to mass import the .csv files for both AUX and National CSV data.

### National Auxillary CSV Data

Please note that the file path for 2023 is slightly different in that the extension is lower case and has an additional folder.

```{r}
#| label: import-AUX-data
#| message: false
#| warning: false
#| echo: true  

acc2007 <- read.csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2008 <- read.csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2009 <- read.csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2010 <- read.csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2011 <- read.csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2012 <- read.csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2013 <- read.csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2014 <- read.csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2015 <- read.csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2016 <- read.csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2017 <- read.csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2018 <- read.csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2019 <- read.csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2020 <- read.csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2021 <- read.csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2022 <- read.csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2023 <- read.csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/acc_aux.csv")
per2007 <- read.csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/PER_AUX.CSV")
per2008 <- read.csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/PER_AUX.CSV")
per2009 <- read.csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/PER_AUX.CSV")
per2010 <- read.csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/PER_AUX.CSV")
per2011 <- read.csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/PER_AUX.CSV")
per2012 <- read.csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/PER_AUX.CSV")
per2013 <- read.csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/PER_AUX.CSV")
per2014 <- read.csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/PER_AUX.CSV")
per2015 <- read.csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/PER_AUX.CSV")
per2016 <- read.csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/PER_AUX.CSV")
per2017 <- read.csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/PER_AUX.CSV")
per2018 <- read.csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/PER_AUX.CSV")
per2019 <- read.csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/PER_AUX.CSV")
per2020 <- read.csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/PER_AUX.CSV")
per2021 <- read.csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/PER_AUX.CSV")
per2022 <- read.csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/PER_AUX.CSV")
per2023 <- read.csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/per_aux.csv")
veh2007 <- read.csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2008 <- read.csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2009 <- read.csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2010 <- read.csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2011 <- read.csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2012 <- read.csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2013 <- read.csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2014 <- read.csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2015 <- read.csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2016 <- read.csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2017 <- read.csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2018 <- read.csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2019 <- read.csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2020 <- read.csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2021 <- read.csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2022 <- read.csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2023 <- read.csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/veh_aux.csv")

```

### NationalCSV

Importing the data files required significant manual effort considering that the naming conventions varied by certain time frames. Please review the following notes to understand the file extension variations:

| Time frame            | File Pathway Notes                                                                         |
|---------------------|---------------------------------------------------|
| 2007-2008, 2023       | Contains extra subfolder                                                                   |
| 2007-2014             | All upper case                                                                             |
| 2015, 2018, 2020-2023 | All lower case                                                                             |
| 2016-2017             | Accident is in lower case and both Person and Vehicle are in proper spelling form.         |
| 2019                  | Accident is in lower case, Person is in proper spelling form, and Vehicle is in lower case |
| 2022-2023             | .csv file extension is in lowercase                                                        |

: NationalCSV Import Notes

```{r}
#| label: import-national-csv-data
#| message: false
#| warning: false
#| echo: true  


accident2007 <- read.csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/ACCIDENT.CSV")
accident2008 <- read.csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/ACCIDENT.CSV")
accident2009 <- read.csv("Raw Data/2009/FARS2009NationalCSV/ACCIDENT.CSV")
accident2010 <- read.csv("Raw Data/2010/FARS2010NationalCSV/ACCIDENT.CSV")
accident2011 <- read.csv("Raw Data/2011/FARS2011NationalCSV/ACCIDENT.CSV")
accident2012 <- read.csv("Raw Data/2012/FARS2012NationalCSV/ACCIDENT.CSV")
accident2013 <- read.csv("Raw Data/2013/FARS2013NationalCSV/ACCIDENT.CSV")
accident2014 <- read.csv("Raw Data/2014/FARS2014NationalCSV/ACCIDENT.CSV")
accident2015 <- read.csv("Raw Data/2015/FARS2015NationalCSV/accident.CSV")
accident2016 <- read.csv("Raw Data/2016/FARS2016NationalCSV/accident.CSV")
accident2017 <- read.csv("Raw Data/2017/FARS2017NationalCSV/accident.CSV")
accident2018 <- read.csv("Raw Data/2018/FARS2018NationalCSV/accident.CSV")
accident2019 <- read.csv("Raw Data/2019/FARS2019NationalCSV/accident.CSV")
accident2020 <- read.csv("Raw Data/2020/FARS2020NationalCSV/accident.CSV")
accident2021 <- read.csv("Raw Data/2021/FARS2021NationalCSV/accident.CSV")
accident2022 <- read.csv("Raw Data/2022/FARS2022NationalCSV/accident.csv")
accident2023 <- read.csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/accident.csv")
person2007 <- read.csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/PERSON.CSV")
person2008 <- read.csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/PERSON.CSV")
person2009 <- read.csv("Raw Data/2009/FARS2009NationalCSV/PERSON.CSV")
person2010 <- read.csv("Raw Data/2010/FARS2010NationalCSV/PERSON.CSV")
person2011 <- read.csv("Raw Data/2011/FARS2011NationalCSV/PERSON.CSV")
person2012 <- read.csv("Raw Data/2012/FARS2012NationalCSV/PERSON.CSV")
person2013 <- read.csv("Raw Data/2013/FARS2013NationalCSV/PERSON.CSV")
person2014 <- read.csv("Raw Data/2014/FARS2014NationalCSV/PERSON.CSV")
person2015 <- read.csv("Raw Data/2015/FARS2015NationalCSV/person.CSV")
person2016 <- read.csv("Raw Data/2016/FARS2016NationalCSV/Person.CSV")
person2017 <- read.csv("Raw Data/2017/FARS2017NationalCSV/Person.CSV")
person2018 <- read.csv("Raw Data/2018/FARS2018NationalCSV/person.CSV")
person2019 <- read.csv("Raw Data/2019/FARS2019NationalCSV/Person.CSV")
person2020 <- read.csv("Raw Data/2020/FARS2020NationalCSV/person.CSV")
person2021 <- read.csv("Raw Data/2021/FARS2021NationalCSV/person.CSV")
person2022 <- read.csv("Raw Data/2022/FARS2022NationalCSV/person.csv")
person2023 <- read.csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/person.csv")
vehicle2007 <- read.csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/VEHICLE.CSV")
vehicle2008 <- read.csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/VEHICLE.CSV")
vehicle2009 <- read.csv("Raw Data/2009/FARS2009NationalCSV/VEHICLE.CSV")
vehicle2010 <- read.csv("Raw Data/2010/FARS2010NationalCSV/VEHICLE.CSV")
vehicle2011 <- read.csv("Raw Data/2011/FARS2011NationalCSV/VEHICLE.CSV")
vehicle2012 <- read.csv("Raw Data/2012/FARS2012NationalCSV/VEHICLE.CSV")
vehicle2013 <- read.csv("Raw Data/2013/FARS2013NationalCSV/VEHICLE.CSV")
vehicle2014 <- read.csv("Raw Data/2014/FARS2014NationalCSV/VEHICLE.CSV")
vehicle2015 <- read.csv("Raw Data/2015/FARS2015NationalCSV/vehicle.CSV")
vehicle2016 <- read.csv("Raw Data/2016/FARS2016NationalCSV/Vehicle.CSV")
vehicle2017 <- read.csv("Raw Data/2017/FARS2017NationalCSV/Vehicle.CSV")
vehicle2018 <- read.csv("Raw Data/2018/FARS2018NationalCSV/vehicle.CSV")
vehicle2019 <- read.csv("Raw Data/2019/FARS2019NationalCSV/vehicle.CSV")
vehicle2020 <- read.csv("Raw Data/2020/FARS2020NationalCSV/vehicle.CSV")
vehicle2021 <- read.csv("Raw Data/2021/FARS2021NationalCSV/vehicle.CSV")
vehicle2022 <- read.csv("Raw Data/2022/FARS2022NationalCSV/vehicle.csv")
vehicle2023 <- read.csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/vehicle.csv")

```

# Data Cleaning

## AUX Stacked Dataframes

-   The ls() function searches for my environment for any object starting with "acc".
-   mget() takes the list of names I defined and gets the actual data

```{r}
#| label: acc-stacked
#| message: false
#| warning: false
#| echo: true  

# define acc data frames from environment
acc_list <- mget(ls(pattern = "^acc20"))
 
# bind data frames together
stacked_acc <- acc_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, FATALS, A_REGION, A_RU)



```

```{r}
#| label: per-stacked
#| message: false
#| warning: false
#| echo: true  

# define per data frames in environment
per_list <- mget(ls(pattern = "^per20"))

# bind data frames
stacked_per <- per_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, VEH_NO, PER_NO, A_PTYPE, A_RESTUSE, A_HELMUSE, A_PERINJ, A_AGE4, A_HRACE)

```

```{r}
#| label: veh-stacked
#| message: false
#| warning: false
#| echo: true  

# define veh data frames in environment
veh_list <- mget(ls(pattern = "^veh20"))

# bind data frames
stacked_veh <- veh_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, VEH_NO)

```

### Create AUX

AUX is the joined data frames of both the stacked ACC AUX files and the stacked PER AUX files.

```{r}
#| label: AUX-join
#| message: false
#| warning: false
#| echo: true 

# Join all AUX files together

AUX <- stacked_per %>% 
  left_join(stacked_acc, by = c("ST_CASE","YEAR"))

```

### Remove old data frames

I want to remove old data frames from my environment to clear up memory and to also make it easier for myself to analyze.

```{r}
#| label: remove-old-df-AUX
#| message: false
#| warning: false
#| echo: true 
#| code-fold: TRUE 

rm(list = ls(pattern = "^acc20")) 
rm(list = ls(pattern = "^per20"))
rm(list = ls(pattern = "^veh20"))
rm(list = ls(pattern = "_list"))


```

## National Stacked Data frames

Note to self: The accident files inherently contain the "year" column.

```{r}
#| label: accident-stacked
#| message: false
#| warning: false
#| echo: true  

# define accident data frames in environment and convert all columns to characters
accident_list <- mget(ls(pattern = "^accident[0-9]")) %>% 
  map(~ .x %>% mutate(across(everything(), as.character)))


# bind data frames
stacked_accident <- accident_list %>% 
  bind_rows(.id = "source_year") %>% 
  select(any_of(c(
    "ST_CASE", "YEAR", "STATE", "COUNTY", "VEH_NO", "PER_NO", 
    "RUR_UR", "RUR_URBNAME", "PERSONS", "FATALS", "A_RU", 
    "longitude", "longitud", "latitude", "LATITUDE", "LONGITUDE", "LONGITUD"
  )))


```

Note to self: The person files inherently DO NOT contain the "year" column. And thus, I will need to manually create said column, and add to each year.

I used AI to help me create a function to add the year column

```{r}
#| label: person-years-function
#| message: false
#| warning: false
#| echo: true  

# Get the names of all objects in your environment that contain a year (4 digits)
person_names <- ls(pattern = "^person\\d{4}")

# Create a named list of these data frames
person_list <- mget(person_names)

# Add year column to each person data frame via function
person_list_updated <- lapply(names(person_list), function(df_name) {
  
  # Extract the year from the string (e.g., "2020")
  year_val <- str_extract(df_name, "\\d{4}")
  
  # Access the data frame
  df <- person_list[[df_name]]
  
  # Add the column
  df$YEAR <- year_val
  
  return(df)
})

# 4. Rename the list elements to match original names
names(person_list_updated) <- person_names

# update R environment with 
list2env(person_list_updated, envir = .GlobalEnv)
```

```{r}
#| label: person-stacked
#| message: false
#| warning: false
#| echo: true  

# define person data frames in environment and convert all columns to characters
person_list <- mget(ls(pattern = "^person[0-9]")) %>% 
  map(~ .x %>% mutate(across(everything(), as.character)))

# bind data frames
stacked_person <- person_list %>% 
  bind_rows(.id = "source_year") %>% 
  select(any_of(c(
    "ST_CASE", "YEAR", "STATE", "COUNTY", "VEH_NO", "PER_NO", 
    "RUR_UR", "RUR_URBNAME", "PERSONS", "FATALS", "A_RU", 
    "longitude", "longitud", "latitude", "LATITUDE", "LONGITUDE", "LONGITUD"
  )))


```

Note to self: The vehicle files DO NOT contain the "year" column. And thus, I will need to manually create said column, and add to each year.

I modified the previous person code chunk for vehicle.

```{r}
#| label: vehicle-years-function
#| message: false
#| warning: false
#| echo: true  

# Get the names of all objects in your environment that contain a year (4 digits)
vehicle_names <- ls(pattern = "^vehicle\\d{4}")

# Create a named list of these data frames
vehicle_list <- mget(vehicle_names)

# Add year column to each vehicle data frame via function
vehicle_list_updated <- lapply(names(vehicle_list), function(df_name) {
  
  # Extract the year from the string (e.g., "2020")
  year_val <- str_extract(df_name, "\\d{4}")
  
  # Access the data frame
  df <- vehicle_list[[df_name]]
  
  # Add the column
  df$YEAR <- year_val
  
  return(df)
})

# 4. Rename the list elements to match original names
names(vehicle_list_updated) <- vehicle_names

# update R environment with 
list2env(vehicle_list_updated, envir = .GlobalEnv)
```

```{r}
#| label: vehicle-stacked
#| message: false
#| warning: false
#| echo: true  

# define vehicle data frames in environment and convert all columns to characters
vehicle_list <- mget(ls(pattern = "^vehicle[0-9]")) %>% 
  map(~ .x %>% mutate(across(everything(), as.character)))

# bind data frames
stacked_vehicle <- vehicle_list %>% 
  bind_rows(.id = "source_year") %>% 
  select(any_of(c(
    "ST_CASE", "YEAR", "STATE", "COUNTY", "VEH_NO", "PER_NO", 
    "RUR_UR", "RUR_URBNAME", "PERSONS", "FATALS", "A_RU", 
    "longitude", "longitud", "latitude", "LATITUDE", "LONGITUDE", "LONGITUD"
  )))

```

### Remove old data frames

I remove obsolete national data frames.

```{r}
#| label: remove-national-df
#| message: false
#| warning: false
#| echo: true  


rm(list = ls(pattern = "accident20"))
rm(list = ls(pattern = "person20"))
rm(list = ls(pattern = "vehicle20"))
rm(list = ls(pattern = "_list"))
rm(list = c("person_names","vehicle_names"))

```

### Create National

National is the joined data frames of the stacked accident, person, and vehicle data frames.

```{r}
#| label: National-join
#| message: false
#| warning: false
#| echo: true 

# Join all national files together

national <- stacked_person %>% 
  left_join(stacked_vehicle, by = c("ST_CASE","YEAR","VEH_NO")) %>% 
  left_join(stacked_accident, by = c("ST_CASE","YEAR")) 

```

# Data Modeling and Visualization

We need to create the following items:

-   A time-series plot that shows the number of deaths by rural vs urban.
    -   We should two lines that represent rural vs urban.
    -   Blue vs red line
    -   y-axis = death percent
-   Also need to have this in tabular form

## Complete Dataset

"df" is defined as the complete dataset combining data from the AUX and NationalCSV datasets.

```{r}
#| label: df-creation
#| message: false
#| warning: false
#| echo: true  

# Make all AUX columns into characters. This makes it easier to join data.
AUX <- AUX %>% 
  map_df(as.character)

df <- AUX %>% 
  left_join(national, by =c("ST_CASE","YEAR","VEH_NO","PER_NO")) %>% 
  select(-ends_with(".y"), -STATE.x)


```

### DF Data Cleaning

Note to self: I need to make sure to use is.na() to find missing data.

```{r}
#| label: df-data-clean
#| message: false
#| warning: false
#| echo: true  

df <- df %>% 
  select(sort(names(.))) %>% ## sort columns alphabetically
  mutate(long = 
           case_when(
             is.na(LONGITUD) ~ longitud, 
             is.na(longitud) ~ LONGITUD,
             .default = longitud
           ),
         lat = 
           case_when(
             is.na(LATITUDE) ~ latitude,
             is.na(latitude) ~ LATITUDE,
             .default = latitude
           ),
         location = 
           case_when(
             RUR_URBNAME.x == "1" ~ "Rural",
             RUR_URBNAME.x == "2" ~ "Urban",
             A_RU == "1" ~ "Rural",
             A_RU == "2" ~ "Urban",
             .default = NA
           )
  ) %>% 
  select(sort(names(.))) %>% ## sort columns alphabetically
  select(ST_CASE, YEAR, PER_NO, VEH_NO, everything())
         

```

### df_v2 subset

I subset the df data frame into df_v2 for data transformation purposes such as removing redundant columns. I keep df in case I need to review the raw data.

```{r}
#| label: df-v2-transform
#| message: false
#| warning: false
#| echo: true  


df_v2 <- df %>% 
  select(-latitude, -LATITUDE, - longitud, -LONGITUD)


```


## Data Export

I export all the datasets so I can upload them to BOX.

```{r}
#| label: label_name
#| message: false
#| warning: false
#| echo: true

# I filter the environment for all data frame objects and create a list
all_data_frames <- mget(ls()) %>% 
  keep(is.data.frame)


# I walk through the entire list and tell R to write .csv files.
iwalk(all_data_frames, ~ write_csv(.x, paste0(.y, ".csv")))

```


Things to do rn

-   Trish will be gone next week (5/12)

    -   she will reach out to me on things that need to be done

-   What to do now?

    -   Create time-series plot and also in tabular form

    -   Create a regression model \[need to meet with Trish to discuss more\]

        -   I should create unadjusted regression model first

            -   This would be a poisson regression or negative binomial

        -   Then create adjusted regression

    -   Make sure to upload files to box

-   send trish a reminder email on monday to send supplmen
