---
title: "NHTSA Data Cleaning"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    fontsize: 0.8em
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 4
    smooth-scroll: TRUE
    embed-resources: TRUE
    fig-align: center
editor: visual
---

# Set-up

## Library

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true  

library(tidyverse)
library(scales)
```

## Import Data

I used Excel to help create functions to mass import the .csv files for both AUX and National CSV data.

### National Auxillary CSV Data

Please note that the file path for 2023 is slightly different in that the extension is lower case and has an additional folder.

```{r}
#| label: import-AUX-data
#| message: false
#| warning: false
#| echo: true  

acc2007 <- read_csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2008 <- read_csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2009 <- read_csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2010 <- read_csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2011 <- read_csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2012 <- read_csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2013 <- read_csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2014 <- read_csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2015 <- read_csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2016 <- read_csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2017 <- read_csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2018 <- read_csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2019 <- read_csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2020 <- read_csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2021 <- read_csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2022 <- read_csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2023 <- read_csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/acc_aux.csv", na = c(""))
per2007 <- read_csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2008 <- read_csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2009 <- read_csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2010 <- read_csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2011 <- read_csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2012 <- read_csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2013 <- read_csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2014 <- read_csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2015 <- read_csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2016 <- read_csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2017 <- read_csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2018 <- read_csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2019 <- read_csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2020 <- read_csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2021 <- read_csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2022 <- read_csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2023 <- read_csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/per_aux.csv", na = c(""))
veh2007 <- read_csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2008 <- read_csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2009 <- read_csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2010 <- read_csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2011 <- read_csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2012 <- read_csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2013 <- read_csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2014 <- read_csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2015 <- read_csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2016 <- read_csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2017 <- read_csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2018 <- read_csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2019 <- read_csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2020 <- read_csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2021 <- read_csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2022 <- read_csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2023 <- read_csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/veh_aux.csv", na = c(""))
```

### NationalCSV

Importing the data files required significant manual effort considering that the naming conventions varied by certain time frames. Please review the following notes to understand the file extension variations:

| Time frame            | File Pathway Notes                                                                         |
|-----------------------|-------------------------------------------------|
| 2007-2008, 2023       | Contains extra subfolder                                                                   |
| 2007-2014             | All upper case                                                                             |
| 2015, 2018, 2020-2023 | All lower case                                                                             |
| 2016-2017             | Accident is in lower case and both Person and Vehicle are in proper spelling form.         |
| 2019                  | Accident is in lower case, Person is in proper spelling form, and Vehicle is in lower case |
| 2022-2023             | .csv file extension is in lowercase                                                        |

: NationalCSV Import Notes

```{r}
#| label: import-national-csv-data
#| message: false
#| warning: false
#| echo: true  


accident2007 <- read_csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/ACCIDENT.CSV")
accident2008 <- read_csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/ACCIDENT.CSV")
accident2009 <- read_csv("Raw Data/2009/FARS2009NationalCSV/ACCIDENT.CSV")
accident2010 <- read_csv("Raw Data/2010/FARS2010NationalCSV/ACCIDENT.CSV")
accident2011 <- read_csv("Raw Data/2011/FARS2011NationalCSV/ACCIDENT.CSV")
accident2012 <- read_csv("Raw Data/2012/FARS2012NationalCSV/ACCIDENT.CSV")
accident2013 <- read_csv("Raw Data/2013/FARS2013NationalCSV/ACCIDENT.CSV")
accident2014 <- read_csv("Raw Data/2014/FARS2014NationalCSV/ACCIDENT.CSV")
accident2015 <- read_csv("Raw Data/2015/FARS2015NationalCSV/accident.CSV")
accident2016 <- read_csv("Raw Data/2016/FARS2016NationalCSV/accident.CSV")
accident2017 <- read_csv("Raw Data/2017/FARS2017NationalCSV/accident.CSV")
accident2018 <- read_csv("Raw Data/2018/FARS2018NationalCSV/accident.CSV")
accident2019 <- read_csv("Raw Data/2019/FARS2019NationalCSV/accident.CSV")
accident2020 <- read_csv("Raw Data/2020/FARS2020NationalCSV/accident.CSV")
accident2021 <- read_csv("Raw Data/2021/FARS2021NationalCSV/accident.CSV")
accident2022 <- read_csv("Raw Data/2022/FARS2022NationalCSV/accident.csv")
accident2023 <- read_csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/accident.csv")
person2007 <- read_csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/PERSON.CSV")
person2008 <- read_csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/PERSON.CSV")
person2009 <- read_csv("Raw Data/2009/FARS2009NationalCSV/PERSON.CSV")
person2010 <- read_csv("Raw Data/2010/FARS2010NationalCSV/PERSON.CSV")
person2011 <- read_csv("Raw Data/2011/FARS2011NationalCSV/PERSON.CSV")
person2012 <- read_csv("Raw Data/2012/FARS2012NationalCSV/PERSON.CSV")
person2013 <- read_csv("Raw Data/2013/FARS2013NationalCSV/PERSON.CSV")
person2014 <- read_csv("Raw Data/2014/FARS2014NationalCSV/PERSON.CSV")
person2015 <- read_csv("Raw Data/2015/FARS2015NationalCSV/person.CSV")
person2016 <- read_csv("Raw Data/2016/FARS2016NationalCSV/Person.CSV")
person2017 <- read_csv("Raw Data/2017/FARS2017NationalCSV/Person.CSV")
person2018 <- read_csv("Raw Data/2018/FARS2018NationalCSV/person.CSV")
person2019 <- read_csv("Raw Data/2019/FARS2019NationalCSV/Person.CSV")
person2020 <- read_csv("Raw Data/2020/FARS2020NationalCSV/person.CSV")
person2021 <- read_csv("Raw Data/2021/FARS2021NationalCSV/person.CSV")
person2022 <- read_csv("Raw Data/2022/FARS2022NationalCSV/person.csv")
person2023 <- read_csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/person.csv")
vehicle2007 <- read_csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/VEHICLE.CSV")
vehicle2008 <- read_csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/VEHICLE.CSV")
vehicle2009 <- read_csv("Raw Data/2009/FARS2009NationalCSV/VEHICLE.CSV")
vehicle2010 <- read_csv("Raw Data/2010/FARS2010NationalCSV/VEHICLE.CSV")
vehicle2011 <- read_csv("Raw Data/2011/FARS2011NationalCSV/VEHICLE.CSV")
vehicle2012 <- read_csv("Raw Data/2012/FARS2012NationalCSV/VEHICLE.CSV")
vehicle2013 <- read_csv("Raw Data/2013/FARS2013NationalCSV/VEHICLE.CSV")
vehicle2014 <- read_csv("Raw Data/2014/FARS2014NationalCSV/VEHICLE.CSV")
vehicle2015 <- read_csv("Raw Data/2015/FARS2015NationalCSV/vehicle.CSV")
vehicle2016 <- read_csv("Raw Data/2016/FARS2016NationalCSV/Vehicle.CSV")
vehicle2017 <- read_csv("Raw Data/2017/FARS2017NationalCSV/Vehicle.CSV")
vehicle2018 <- read_csv("Raw Data/2018/FARS2018NationalCSV/vehicle.CSV")
vehicle2019 <- read_csv("Raw Data/2019/FARS2019NationalCSV/vehicle.CSV")
vehicle2020 <- read_csv("Raw Data/2020/FARS2020NationalCSV/vehicle.CSV")
vehicle2021 <- read_csv("Raw Data/2021/FARS2021NationalCSV/vehicle.CSV")
vehicle2022 <- read_csv("Raw Data/2022/FARS2022NationalCSV/vehicle.csv")
vehicle2023 <- read_csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/vehicle.csv")

```

# Data Cleaning

## AUX Stacked Dataframes

-   The ls() function searches for my environment for any object starting with "acc".
-   mget() takes the list of names I defined and gets the actual data

```{r}
#| label: acc-stacked
#| message: false
#| warning: false
#| echo: true  

# define acc data frames from environment
acc_list <- mget(ls(pattern = "^acc20"))
 
# bind data frames together
stacked_acc <- acc_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, FATALS, A_REGION, A_RU)



```

```{r}
#| label: per-stacked
#| message: false
#| warning: false
#| echo: true  

# define per data frames in environment
per_list <- mget(ls(pattern = "^per20"))

# bind data frames
stacked_per <- per_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, VEH_NO, PER_NO, A_PTYPE, A_RESTUSE, A_HELMUSE, A_PERINJ, A_AGE4, A_HRACE)

```

```{r}
#| label: veh-stacked
#| message: false
#| warning: false
#| echo: true  

# define veh data frames in environment
veh_list <- mget(ls(pattern = "^veh20"))

# bind data frames
stacked_veh <- veh_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, VEH_NO)

```

### Create AUX

AUX is the joined data frames of both the stacked ACC AUX files and the stacked PER AUX files.

```{r}
#| label: AUX-join
#| message: false
#| warning: false
#| echo: true 

# Join all AUX files together

AUX <- stacked_per %>% 
  left_join(stacked_acc, by = c("ST_CASE","YEAR"))

```

### Remove old data frames

I want to remove old data frames from my environment to clear up memory and to also make it easier for myself to analyze.

```{r}
#| label: remove-old-df-AUX
#| message: false
#| warning: false
#| echo: true 
#| code-fold: TRUE 

rm(list = ls(pattern = "^acc20")) 
rm(list = ls(pattern = "^per20"))
rm(list = ls(pattern = "^veh20"))
rm(list = ls(pattern = "_list"))


```

## National Stacked Data frames

Note to self: The accident files inherently contain the "year" column.

```{r}
#| label: accident-stacked
#| message: false
#| warning: false
#| echo: true  

# define accident data frames in environment and convert all columns to characters
accident_list <- mget(ls(pattern = "^accident[0-9]")) %>% 
  map(~ .x %>% mutate(across(everything(), as.character)))


# bind data frames
stacked_accident <- accident_list %>% 
  bind_rows(.id = "source_year") %>% 
  select(any_of(c(
    "ST_CASE", 
    "YEAR", 
    "STATE", 
    "COUNTY", 
    "VEH_NO", 
    "PER_NO", 
    "PERSONS", 
### Geographic Indicators
    "RUR_URB", 
    "RUR_URBNAME",
    "A_RU",
    "longitude", 
    "longitud", 
    "latitude", 
    "LATITUDE", 
    "LONGITUDE", 
    "LONGITUD", 
  "FUNC_SYS 1",
  "FUNC_SYS 2",
  "FUNC_SYS 3",
  "FUNC_SYS 4",
  "FUNC_SYS 5",
  "FUNC_SYS 6",
  "FUNC_SYS 7",
"ROAD_FNC",
### Time of Accident
  "MONTH",
  "MONTHNAME",
  "DAY",
  "DAYNAME",
  "DAY_WEEK",
  "DAY_WEEKNAME",
  "YEAR",
  "HOUR",
  "MINUTE",
### EMS variables
"NOT_HOUR",
"NOT_MIN",
"ARR_HOUR",
"ARR_MIN",
"HOSP_HR",
"HOSP_MN",
"FATALS"

  )))


```

Note to self: The person files inherently DO NOT contain the "year" column. And thus, I will need to manually create said column, and add to each year.

I used AI to help me create a function to add the year column

```{r}
#| label: person-years-function
#| message: false
#| warning: false
#| echo: true  

# Get the names of all objects in your environment that contain a year (4 digits)
person_names <- ls(pattern = "^person\\d{4}")

# Create a named list of these data frames
person_list <- mget(person_names)

# Add year column to each person data frame via function
person_list_updated <- lapply(names(person_list), function(df_name) {
  
  # Extract the year from the string (e.g., "2020")
  year_val <- str_extract(df_name, "\\d{4}")
  
  # Access the data frame
  df <- person_list[[df_name]]
  
  # Add the column
  df$YEAR <- year_val
  
  return(df)
})

# 4. Rename the list elements to match original names
names(person_list_updated) <- person_names

# update R environment with 
list2env(person_list_updated, envir = .GlobalEnv)
```

```{r}
#| label: person-stacked
#| message: false
#| warning: false
#| echo: true  

# define person data frames in environment and convert all columns to characters
person_list <- mget(ls(pattern = "^person[0-9]")) %>% 
  map(~ .x %>% mutate(across(everything(), as.character)))

# bind data frames
stacked_person <- person_list %>% 
  bind_rows(.id = "source_year") %>% 
  select(any_of(c(
    "ST_CASE", 
    "YEAR", 
    "STATE", 
    "COUNTY", 
    "VEH_NO", 
    "PER_NO", 
    "RUR_URB", 
    "RUR_URBNAME", 
    "PERSONS", 
    "FATALS", 
    "A_RU", 
    "longitude", 
    "longitud", 
    "latitude", 
    "LATITUDE", 
    "LONGITUDE", 
    "LONGITUD",
### Demographics
    "AGE",
    "SEX",
### behavioral variables
    "ALC_STATUS",
    "ALC_RES",
    "DRUGS",
    "DRUGSNAME",
    "DSTATUS",
    "REST_USE",
    "REST_MIS",
    "REST_USENAME",
    "HELM_USE",
    "HELM_USENAME",
### EMS variables
"INJ_SEV",
"INJ_SEVNAME",
"HOSPITAL",
"HOSPITALNAME",
"DOA",
"DOANAME",
"DEATH_DA",
"DEATH_YR",
"LAG_HRS",
"LAG_MINS",
"BODY_TYP",
"BODY_TYPNAME"
  )))


```

Note to self: The vehicle files DO NOT contain the "year" column. And thus, I will need to manually create said column, and add to each year.

I modified the previous person code chunk for vehicle.

```{r}
#| label: vehicle-years-function
#| message: false
#| warning: false
#| echo: true  

# Get the names of all objects in your environment that contain a year (4 digits)
vehicle_names <- ls(pattern = "^vehicle\\d{4}")

# Create a named list of these data frames
vehicle_list <- mget(vehicle_names)

# Add year column to each vehicle data frame via function
vehicle_list_updated <- lapply(names(vehicle_list), function(df_name) {
  
  # Extract the year from the string (e.g., "2020")
  year_val <- str_extract(df_name, "\\d{4}")
  
  # Access the data frame
  df <- vehicle_list[[df_name]]
  
  # Add the column
  df$YEAR <- year_val
  
  return(df)
})

# 4. Rename the list elements to match original names
names(vehicle_list_updated) <- vehicle_names

# update R environment with 
list2env(vehicle_list_updated, envir = .GlobalEnv)
```

```{r}
#| label: vehicle-stacked
#| message: false
#| warning: false
#| echo: true  

# define vehicle data frames in environment and convert all columns to characters
vehicle_list <- mget(ls(pattern = "^vehicle[0-9]")) %>% 
  map(~ .x %>% mutate(across(everything(), as.character)))

# bind data frames
stacked_vehicle <- vehicle_list %>% 
  bind_rows(.id = "source_year") %>% 
  select(any_of(c(
    "ST_CASE", 
    "YEAR", 
    "STATE", 
    "COUNTY", 
    "VEH_NO", 
    "PER_NO", 
    "RUR_URB", 
    "RUR_URBNAME", 
    "PERSONS", 
    "FATALS", 
    "A_RU",
    "longitude", 
    "longitud", 
    "latitude", 
    "LATITUDE", 
    "LONGITUDE", 
    "LONGITUD",
### behavioral variables
    "DR_DRINK",
    "SPEEDREL",
    "PREV_SPD",
"DR_ZIP",
"SPEC_USE"
  )))

```

### Remove old data frames

I remove obsolete national data frames.

```{r}
#| label: remove-national-df
#| message: false
#| warning: false
#| echo: true  


rm(list = ls(pattern = "accident20"))
rm(list = ls(pattern = "person20"))
rm(list = ls(pattern = "vehicle20"))
rm(list = ls(pattern = "_list"))
rm(list = c("person_names","vehicle_names"))

```

### Create National

National is the joined data frames of the stacked accident, person, and vehicle data frames.

```{r}
#| label: National-join
#| message: false
#| warning: false
#| echo: true 

# Join all national files together

national <- stacked_person %>% 
  left_join(stacked_vehicle, by = c("ST_CASE","YEAR","VEH_NO")) %>% 
  left_join(stacked_accident, by = c("ST_CASE","YEAR")) 

```

# Data Modeling and Visualization

We need to create the following items:

-   A time-series plot that shows the number of deaths by rural vs urban.
    -   We should two lines that represent rural vs urban.
    -   Blue vs red line
    -   y-axis = death percent
-   Also need to have this in tabular form

## Complete Dataset

"df" is defined as the complete dataset combining data from the AUX and NationalCSV datasets.

```{r}
#| label: df-creation
#| message: false
#| warning: false
#| echo: true  

# Make all AUX columns into characters. This makes it easier to join data.
AUX <- AUX %>% 
  map_df(as.character)

df <- AUX %>% 
  left_join(national, by =c("ST_CASE","YEAR","VEH_NO","PER_NO")) %>% 
  select(-ends_with(".y"), -STATE.x)


```

### DF Data Cleaning

Note to self: I need to make sure to use is.na() to find missing data.

```{r}
#| label: df-data-clean
#| message: false
#| warning: false
#| echo: true  

df <- df %>% 
  select(sort(names(.))) %>% ## sort columns alphabetically
  mutate(long = 
           case_when(
             is.na(LONGITUD) ~ longitud, 
             is.na(longitud) ~ LONGITUD,
             .default = longitud
           ),
         lat = 
           case_when(
             is.na(LATITUDE) ~ latitude,
             is.na(latitude) ~ LATITUDE,
             .default = latitude
           ),
         location = 
           case_when(
             RUR_URBNAME.x == "1" ~ "Rural",
             RUR_URBNAME.x == "2" ~ "Urban",
             A_RU == "1" ~ "Rural",
             A_RU == "2" ~ "Urban",
             .default = NA
           )
  ) %>% 
  select(sort(names(.))) %>% ## sort columns alphabetically
  select(ST_CASE, YEAR, PER_NO, VEH_NO, everything())
         

```

### df_v2 subset

I subset the df data frame into df_v2 for data transformation purposes such as removing redundant columns. I keep df in case I need to review the raw data.

```{r}
#| label: df-v2-transform
#| message: false
#| warning: false
#| echo: true  


df_v2 <- df %>% 
  select(-latitude, -LATITUDE, - longitud, -LONGITUD)


```

## Data Export

I export all the datasets so I can upload them to BOX.

```{r}
#| label: AUX_rename
#| message: false
#| warning: false
#| echo: true  

# I need to rewrite this to prevent errors
AUX_data <- AUX
rm(AUX)

```

```{r}
#| label: data_export
#| message: false
#| warning: false
#| echo: true

# I filter the environment for all data frame objects and create a list
all_data_frames <- mget(ls()) %>% 
  keep(is.data.frame)


# I walk through the entire list and tell R to write .csv files into the "Export" folder.
iwalk(all_data_frames, ~ write_csv(.x, paste0("Exports/", .y, ".csv")))



```

Things to do rn

-   Trish will be gone next week (5/12)

    -   she will reach out to me on things that need to be done

-   What to do now?

    -   Create time-series plot and also in tabular form

    -   Create a regression model \[need to meet with Trish to discuss more\]

        -   I should create unadjusted regression model first

            -   This would be a poisson regression or negative binomial

        -   Then create adjusted regression

    -   Make sure to upload files to box

-   send trish a reminder email on monday to send supplmen

# Troubleshoot Data

## General Overview + Thoughts

-   I want to troubleshoot the NAs by investigating INJ_SEV NA values in the 2015 person dataset. I need to find out which step in data loading results in these NA values in being created.

    -   The other variables listed also generally contain NA values.
    -   Okay so by running colSums(is.na()) I've found a buttload of NA values for the aforementioned variables. This def confirms that something was wrong when I loaded in data

    ```{r}
    #| label: troubleshoot_df_v2
    #| message: false
    #| warning: false
    #| echo: true  

    df_v2_troubleshoot <- df_v2 %>% 
      select(ST_CASE, YEAR, INJ_SEV,INJ_SEVNAME, BODY_TYP, SPEC_USE)

    colSums(is.na(df_v2_troubleshoot))

    ```

-   Thought: Idk if something is wrong w the entire dataset or it is specific years like 2015, 2021, and 2022. Thus I will compare 2014 against 2015 to see if there are NA values.

    -   okay, so it looks like after comparing the year NA values - it looks like this is an overall loading error since NA values are appearing in both years.

```{r}
#| label: compare_2014_vs_2015_na
#| message: false
#| warning: false
#| echo: true  

na_2014 <- df_v2 %>% 
  select(ST_CASE, YEAR, INJ_SEV,INJ_SEVNAME, BODY_TYP, SPEC_USE) %>% 
  filter(YEAR == 2014)

colSums(is.na(na_2014))

na_2015 <- df_v2 %>% 
  select(ST_CASE, YEAR, INJ_SEV,INJ_SEVNAME, BODY_TYP, SPEC_USE) %>% 
  filter(YEAR == 2015)

colSums(is.na(na_2015))



```

## stacked_person analysis

-   I am going to check if something wonky happened when I stacked the person files together.
    -   Note: This does not verify the SPEC_USE variable since this variable is from a stacked_vehicle
    -   okay this is interesting. It looks like we only have 1 INJ_SEV NA value and a buttload of BODY_TYP NA values.
        -   Therefore, this means that something went wrong when the files merged. One problem at a time. Need to solve the NA INJ_SEV.

```{r}
#| label: troubleshoot_stacked_person
#| message: false
#| warning: false
#| echo: true  

troubleshoot_stacked_person <- stacked_person %>% 
  select(ST_CASE, YEAR, INJ_SEV,INJ_SEVNAME, BODY_TYP)

colSums(is.na(troubleshoot_stacked_person))

view(troubleshoot_stacked_person)
```

-   In this code chunk, I am isolating the NA value to find out more information
    -   Okay weird, it just shows me ST_CASE = 10109 has a singular NA value.

```{r}
#| label: troubleshoot_stacked_person_NA
#| message: false
#| warning: false
#| echo: true  

troubleshoot_stacked_person_NA <- stacked_person %>% 
  select(ST_CASE, YEAR, INJ_SEV,INJ_SEVNAME, BODY_TYP) %>% 
  filter(is.na(INJ_SEV))

colSums(is.na(troubleshoot_stacked_person_NA))

view(troubleshoot_stacked_person_NA)
```

-   I'm going to check out ST_CASE 10109 to see if there are duplicates or something.
    -   Okay this is really weird. The raw data is the source of truth. I am expecting 3 rows where ST_CASE = 10109 and YEAR == 2015. Instead I get a singular NA row when YEAR == 2015.
        -   This means I need to go further back and check how the data is being loaded in.

```{r}
#| label: troubleshoot_stacked_person_NA
#| message: false
#| warning: false
#| echo: true  

troubleshoot_stacked_person_10109 <- stacked_person %>% 
  select(ST_CASE, YEAR, INJ_SEV,INJ_SEVNAME, BODY_TYP) %>% 
  filter(ST_CASE == 10109)

colSums(is.na(troubleshoot_stacked_person_10109))

view(troubleshoot_stacked_person_10109)
```

### stacked_person_2015 analysis

#### INJ_SEV FIX

-   Perhaps the way R is parsing data files is incorrect. I'm going to try and force R to remove any invisible spaces before it reads data to see if this fixes the issue. I'm doing this by converting read.csv() to read_csv().
    -   I'm also going to force R to read everything as a character
-   **Wait I actually found the issue. Looking at the Raw excel file, we have 81621 rows from the 2015 person files. HOWEVER, R is only reading in 256 of those rows. I need to figure out how to fix this issue.**
    -   **According to Gemini, I should use read_csv instead of read_csv. The latter is old school and is less flexible to weird characters.**

```{r}
#| label: troubleshoot_stacked_person_analysis_v2
#| message: false
#| warning: false
#| echo: true  

# load data in again
person2015_v2 <- read_csv("Raw Data/2015/FARS2015NationalCSV/person.CSV")

# analyze file

person_2015_troubleshoot_v2 <- person2015_v2 %>% 
  select(ST_CASE, INJ_SEV, BODY_TYP)

colSums(is.na(person_2015_troubleshoot_v2))


view(person_2015_troubleshoot_v2)
```

#### BODY_TYP FIX

-   Okay so the INJ_SEV is solved but I need to figure out why BODY_TYP is still flipping out

```{r}
#| label: BODY_TYP_FIX
#| message: false
#| warning: false
#| echo: true  

# analyze file

person_troubleshoot_BODY_FIX <- stacked_person %>% 
  select(ST_CASE,YEAR, INJ_SEV, BODY_TYP) %>% 
  filter(is.na(BODY_TYP))

colSums(is.na(person_troubleshoot_BODY_FIX))


view(person_troubleshoot_BODY_FIX)

```

-   I filter to the year 2007 and ST_CASE == 10015 to see if it loaded in things correctly.
-   Interesting. Things did load correctly. Perhaps this means that all these NAs are actually correct?

```{r}
#| label: BODY_TYP_FIX_2007
#| message: false
#| warning: false
#| echo: true  

# analyze file

person_troubleshoot_BODY_FIX_2007 <- stacked_person %>% 
  select(ST_CASE,YEAR, INJ_SEV, BODY_TYP) %>% 
  filter(YEAR == 2007,
         ST_CASE == 10015)

colSums(is.na(person_troubleshoot_BODY_FIX_2007))


view(person_troubleshoot_BODY_FIX_2007)

```

### Troubleshoot Conclusion and Sanity Check

-   I think everything was fixed by changing read.csv to read_csv. To ensure that there are no false NA values (e.g R reads it as NA but there is actual data in the source file), I will randomly select 5 NA BODY_TYP rows + 5 NA SPEC_USE rows and verify that they are correct (e.g they match up w the raw data in Excel.)

-   

    ## The issue is R's parsing abilities. R reads only the first hundred lines of code and uses that to guess the next hundred.

| ST_CASE | YEAR | BODY_TYP Verified |
|---------|------|-------------------|
| 10015   | 2007 | 1                 |
| 210430  | 2015 | 1                 |
| 210452  | 2020 | 1                 |
| 530467  | 2023 | 1                 |
| 60390   | 2010 | 1                 |

```{r}
#| label: SPEC_USE_NA
#| message: false
#| warning: false
#| echo: true  


SPEC_USE_NA <- df_v2 %>% 
  select(ST_CASE, YEAR, SPEC_USE) %>% 
  filter(is.na(SPEC_USE))

view(SPEC_USE_NA)

```

-   Trying to diagnose encoding issue.

    -   According to ChatGPT, it's not a data conversion issue but rather an encoding one. How R

```{r}
#| label: SPEC_USE_2018
#| message: false
#| warning: false
#| echo: true  

vehicle2018 <- read_csv("Raw Data/2018/FARS2018NationalCSV/vehicle.CSV", 
                        col_types = cols(.default = col_character()))

## Hm... This is not a parsing issue since I set everything to a character.

```

-   I'm trying to figure out where the duplicates occur.

```{r}
#| label: SPEC_USE_2018_v2
#| message: false
#| warning: false
#| echo: true  

vehicle2018_v2 <- df_v2 %>% 
  filter(YEAR == 2018 & 
         ST_CASE == 131083)

view(vehicle2018_v2)

```

-   Looks like the duplicates don't occur when we stack the vehicle data together. **The duplicates occur when we merge the data!!!!**

```{r}
#| label: SPEC_USE_2018_v3
#| message: false
#| warning: false
#| echo: true  

vehicle2018_v3 <- stacked_vehicle %>%
  filter(YEAR == 2018 & 
         ST_CASE == 131083) 

view(vehicle2018_v3)
```

-   Perhaps the duplicates occur in the first merge?

    -   **BINGO!!! The duplicates do occur in the first merge!!!**

```{r}
#| label: SPEC_USE_2018_v4
#| message: false
#| warning: false
#| echo: true  

vehicle2018_v4 <- national %>%
  filter(YEAR == 2018 & 
         ST_CASE == 131083) 

view(vehicle2018_v4)
```

| ST_CASE | YEAR | SPEC_USE Verified                                                                    |
|---------|------|--------------------------------------------------------------------------------------|
| 131083  | 2018 | 0 \<-- **need to correct. This is not accurate. There is exactly 1 row not 6 rows.** |
| 480376  | 2022 |                                                                                      |
| 560119  | 2023 |                                                                                      |
| 62292   | 2008 |                                                                                      |
| 510541  | 2010 | 1                                                                                    |

-   I also conducted a meta analysis to see if the percentage of NAs makes sense

```{r}
#| label: tbl-df_NA_BODY_TYP
#| tbl-cap: "df BODY_TYP Meta Analysis"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true


df_NA_BODY_TYP <- df_v2 %>% 
  select(ST_CASE, YEAR, INJ_SEV, BODY_TYP, SPEC_USE) %>% 
  group_by(YEAR) %>% 
  summarize(
    BODY_TYP_sum = sum(!is.na(BODY_TYP)),
    BODY_TYP_NA = sum(is.na(BODY_TYP))) %>% 
  mutate(BODY_TYP_NA_percent = percent(BODY_TYP_NA / BODY_TYP_sum, accuracy = 0.01))

df_NA_BODY_TYP

```

-   I did the same thing with SPEC_USE

```{r}
#| label: tbl-df_NA_SPEC_USE
#| tbl-cap: "df SPEC_USE Meta Analysis"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true


df_NA_SPEC_USE <- df_v2 %>% 
  select(ST_CASE, YEAR, INJ_SEV, SPEC_USE, SPEC_USE) %>% 
  group_by(YEAR) %>% 
  summarize(
    SPEC_USE_sum = sum(!is.na(SPEC_USE)),
    SPEC_USE_NA = sum(is.na(SPEC_USE))) %>% 
  mutate(SPEC_USE_NA_percent = percent(SPEC_USE_NA / SPEC_USE_sum, accuracy = 0.01))

df_NA_SPEC_USE

```

-   Now to really check that everything is okay, I'm going to compare these results to a variable we haven't had an issue with previously. I'll use DR_ZIP as an example
    -   Okay this is good. DR_ZIP is showing similar NA percentage values.

```{r}
#| label: tbl-df_NA_DR_ZIP
#| tbl-cap: "df DR_ZIP Meta Analysis"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true


df_NA_DR_ZIP <- df_v2 %>% 
  select(ST_CASE, YEAR, INJ_SEV, DR_ZIP, DR_ZIP) %>% 
  group_by(YEAR) %>% 
  summarize(
    DR_ZIP_sum = sum(!is.na(DR_ZIP)),
    DR_ZIP_NA = sum(is.na(DR_ZIP))) %>% 
  mutate(DR_ZIP_NA_percent = percent(DR_ZIP_NA / DR_ZIP_sum, accuracy = 0.01))

df_NA_DR_ZIP

```
