---
title: "NHTSA Data Cleaning"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    fontsize: 0.8em
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 4
    smooth-scroll: TRUE
    embed-resources: TRUE
    fig-align: center
editor: visual
---

# Set-up

## Library

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true  

library(tidyverse)
```

## Import Data

I used Excel to help create functions to mass import the .csv files for both AUX and National CSV data.

### National Auxillary CSV Data

Please note that the file path for 2023 is slightly different in that the extension is lower case and has an additional folder.

```{r}
#| label: import-AUX-data
#| message: false
#| warning: false
#| echo: true  

acc2007 <- read.csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2008 <- read.csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2009 <- read.csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2010 <- read.csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2011 <- read.csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2012 <- read.csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2013 <- read.csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2014 <- read.csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2015 <- read.csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2016 <- read.csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2017 <- read.csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2018 <- read.csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2019 <- read.csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2020 <- read.csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2021 <- read.csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2022 <- read.csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/ACC_AUX.CSV")
acc2023 <- read.csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/acc_aux.csv")
per2007 <- read.csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/PER_AUX.CSV")
per2008 <- read.csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/PER_AUX.CSV")
per2009 <- read.csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/PER_AUX.CSV")
per2010 <- read.csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/PER_AUX.CSV")
per2011 <- read.csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/PER_AUX.CSV")
per2012 <- read.csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/PER_AUX.CSV")
per2013 <- read.csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/PER_AUX.CSV")
per2014 <- read.csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/PER_AUX.CSV")
per2015 <- read.csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/PER_AUX.CSV")
per2016 <- read.csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/PER_AUX.CSV")
per2017 <- read.csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/PER_AUX.CSV")
per2018 <- read.csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/PER_AUX.CSV")
per2019 <- read.csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/PER_AUX.CSV")
per2020 <- read.csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/PER_AUX.CSV")
per2021 <- read.csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/PER_AUX.CSV")
per2022 <- read.csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/PER_AUX.CSV")
per2023 <- read.csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/per_aux.csv")
veh2007 <- read.csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2008 <- read.csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2009 <- read.csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2010 <- read.csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2011 <- read.csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2012 <- read.csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2013 <- read.csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2014 <- read.csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2015 <- read.csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2016 <- read.csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2017 <- read.csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2018 <- read.csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2019 <- read.csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2020 <- read.csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2021 <- read.csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2022 <- read.csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/VEH_AUX.CSV")
veh2023 <- read.csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/veh_aux.csv")

```

### NationalCSV

Importing the data files required significant manual effort considering that the naming conventions varied by certain time frames. Please review the following notes to understand the file extension variations:

| Time frame            | File Pathway Notes                                                                          |
|-----------------------|---------------------------------------------------------------------------------------------|
| 2007-2008, 2022, 2023 | Contains extra subfolder                                                                    |
| 2007-2014             | All upper case                                                                              |
| 2015, 2018, 2020-2023 | All lower case                                                                              |
| 2016-2017             | Accident is in lower case and both Person and Vehicle are in proper spelling form.          |
| 2019                  | Accident is in lower case, Person is in proper spelling form, and Vehicle is in lower case  |
| 2022-2023             | .csv file extension is in lowercase                                                         |

: NationalCSV Import Notes

```{r}
#| label: import-national-csv-data
#| message: false
#| warning: false
#| echo: true  

accident2007 <- read.csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/ACCIDENT.CSV")
accident2008 <- read.csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/ACCIDENT.CSV")
accident2009 <- read.csv("Raw Data/2009/FARS2009NationalCSV/ACCIDENT.CSV")
accident2010 <- read.csv("Raw Data/2010/FARS2010NationalCSV/ACCIDENT.CSV")
accident2011 <- read.csv("Raw Data/2011/FARS2011NationalCSV/ACCIDENT.CSV")
accident2012 <- read.csv("Raw Data/2012/FARS2012NationalCSV/ACCIDENT.CSV")
accident2013 <- read.csv("Raw Data/2013/FARS2013NationalCSV/ACCIDENT.CSV")
accident2014 <- read.csv("Raw Data/2014/FARS2014NationalCSV/ACCIDENT.CSV")
accident2015 <- read.csv("Raw Data/2015/FARS2015NationalCSV/accident.CSV")
accident2016 <- read.csv("Raw Data/2016/FARS2016NationalCSV/accident.CSV")
accident2017 <- read.csv("Raw Data/2017/FARS2017NationalCSV/accident.CSV")
accident2018 <- read.csv("Raw Data/2018/FARS2018NationalCSV/accident.CSV")
accident2019 <- read.csv("Raw Data/2019/FARS2019NationalCSV/accident.CSV")
accident2020 <- read.csv("Raw Data/2020/FARS2020NationalCSV/accident.CSV")
accident2021 <- read.csv("Raw Data/2021/FARS2021NationalCSV/accident.CSV")
accident2022 <- read.csv("Raw Data/2022/FARS2022NationalCSV/FARS2022NationalCSV/accident.CSV")
accident2023 <- read.csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/accident.CSV")
person2007 <- read.csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/PERSON.CSV")
person2008 <- read.csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/PERSON.CSV")
person2009 <- read.csv("Raw Data/2009/FARS2009NationalCSV/PERSON.CSV")
person2010 <- read.csv("Raw Data/2010/FARS2010NationalCSV/PERSON.CSV")
person2011 <- read.csv("Raw Data/2011/FARS2011NationalCSV/PERSON.CSV")
person2012 <- read.csv("Raw Data/2012/FARS2012NationalCSV/PERSON.CSV")
person2013 <- read.csv("Raw Data/2013/FARS2013NationalCSV/PERSON.CSV")
person2014 <- read.csv("Raw Data/2014/FARS2014NationalCSV/PERSON.CSV")
person2015 <- read.csv("Raw Data/2015/FARS2015NationalCSV/person.CSV")
person2016 <- read.csv("Raw Data/2016/FARS2016NationalCSV/Person.CSV")
person2017 <- read.csv("Raw Data/2017/FARS2017NationalCSV/Person.CSV")
person2018 <- read.csv("Raw Data/2018/FARS2018NationalCSV/person.CSV")
person2019 <- read.csv("Raw Data/2019/FARS2019NationalCSV/Person.CSV")
person2020 <- read.csv("Raw Data/2020/FARS2020NationalCSV/person.CSV")
person2021 <- read.csv("Raw Data/2021/FARS2021NationalCSV/person.CSV")
person2022 <- read.csv("Raw Data/2022/FARS2022NationalCSV/FARS2022NationalCSV/person.CSV")
person2023 <- read.csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/person.CSV")
vehicle2007 <- read.csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/VEHICLE.CSV")
vehicle2008 <- read.csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/VEHICLE.CSV")
vehicle2009 <- read.csv("Raw Data/2009/FARS2009NationalCSV/VEHICLE.CSV")
vehicle2010 <- read.csv("Raw Data/2010/FARS2010NationalCSV/VEHICLE.CSV")
vehicle2011 <- read.csv("Raw Data/2011/FARS2011NationalCSV/VEHICLE.CSV")
vehicle2012 <- read.csv("Raw Data/2012/FARS2012NationalCSV/VEHICLE.CSV")
vehicle2013 <- read.csv("Raw Data/2013/FARS2013NationalCSV/VEHICLE.CSV")
vehicle2014 <- read.csv("Raw Data/2014/FARS2014NationalCSV/VEHICLE.CSV")
vehicle2015 <- read.csv("Raw Data/2015/FARS2015NationalCSV/vehicle.CSV")
vehicle2016 <- read.csv("Raw Data/2016/FARS2016NationalCSV/Vehicle.CSV")
vehicle2017 <- read.csv("Raw Data/2017/FARS2017NationalCSV/Vehicle.CSV")
vehicle2018 <- read.csv("Raw Data/2018/FARS2018NationalCSV/vehicle.CSV")
vehicle2019 <- read.csv("Raw Data/2019/FARS2019NationalCSV/vehicle.CSV")
vehicle2020 <- read.csv("Raw Data/2020/FARS2020NationalCSV/vehicle.CSV")
vehicle2021 <- read.csv("Raw Data/2021/FARS2021NationalCSV/vehicle.CSV")
vehicle2022 <- read.csv("Raw Data/2022/FARS2022NationalCSV/FARS2022NationalCSV/vehicle.CSV")
vehicle2023 <- read.csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/vehicle.CSV")


```

# Data Cleaning

## AUX Stacked Dataframes

-   The ls() function searches for my environment for any object starting with "acc".
-   mget() takes the list of names I defined and gets the actual data

```{r}
#| label: acc-stacked
#| message: false
#| warning: false
#| echo: true  

# define acc data frames from environment
acc_list <- mget(ls(pattern = "^acc"))
 
# bind data frames together
stacked_acc <- acc_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, FATALS, A_REGION, A_RU)



```

```{r}
#| label: per-stacked
#| message: false
#| warning: false
#| echo: true  

# define per data frames in environment
per_list <- mget(ls(pattern = "^per"))

# bind data frames
stacked_per <- per_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, VEH_NO, PER_NO, A_PTYPE, A_RESTUSE, A_HELMUSE, A_PERINJ, A_AGE4, A_HRACE)

```

```{r}
#| label: veh-stacked
#| message: false
#| warning: false
#| echo: true  

# define veh data frames in environment
veh_list <- mget(ls(pattern = "^veh"))

# bind data frames
stacked_veh <- veh_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, VEH_NO)

```

### Create AUX

AUX is the joined data frames of both the stacked ACC AUX files and the stacked PER AUX files.

```{r}
#| label: AUX-join
#| message: false
#| warning: false
#| echo: true 

# Join all AUX files together

AUX <- stacked_per %>% 
  left_join(stacked_acc, by = c("ST_CASE","YEAR"))

```

## National Stacked Data frames

## Remove old data frames

I want to remove old data frames from my environment to clear up memory and to also make it easier for myself to analyze.

```{r}
#| label: remove-old-df
#| message: false
#| warning: false
#| echo: true 
#| code-fold: TRUE 

rm(list = ls(pattern = "^acc")) 
rm(list = ls(pattern = "^per"))
rm(list = ls(pattern = "^veh"))


```
