---
title: "NHTSA Data Cleaning"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    fontsize: 0.8em
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 4
    smooth-scroll: TRUE
    embed-resources: TRUE
    fig-align: center
editor: visual
---

# Set-up

## Library

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true  

library(tidyverse)
library(scales)
```

## Import Data

I used Excel to help create functions to mass import the .csv files for both AUX and National CSV data.

### National Auxillary CSV Data

Please note that the file path for 2023 is slightly different in that the extension is lower case and has an additional folder.

```{r}
#| label: import-AUX-data
#| message: false
#| warning: false
#| echo: true  

acc2007 <- read_csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2008 <- read_csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2009 <- read_csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2010 <- read_csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2011 <- read_csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2012 <- read_csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2013 <- read_csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2014 <- read_csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2015 <- read_csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2016 <- read_csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2017 <- read_csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2018 <- read_csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2019 <- read_csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2020 <- read_csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2021 <- read_csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2022 <- read_csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/ACC_AUX.CSV", na = c(""))
acc2023 <- read_csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/acc_aux.csv", na = c(""))
per2007 <- read_csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2008 <- read_csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2009 <- read_csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2010 <- read_csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2011 <- read_csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2012 <- read_csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2013 <- read_csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2014 <- read_csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2015 <- read_csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2016 <- read_csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2017 <- read_csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2018 <- read_csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2019 <- read_csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2020 <- read_csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2021 <- read_csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2022 <- read_csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/PER_AUX.CSV", na = c(""))
per2023 <- read_csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/per_aux.csv", na = c(""))
veh2007 <- read_csv("Raw Data/2007/FARS2007NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2008 <- read_csv("Raw Data/2008/FARS2008NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2009 <- read_csv("Raw Data/2009/FARS2009NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2010 <- read_csv("Raw Data/2010/FARS2010NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2011 <- read_csv("Raw Data/2011/FARS2011NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2012 <- read_csv("Raw Data/2012/FARS2012NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2013 <- read_csv("Raw Data/2013/FARS2013NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2014 <- read_csv("Raw Data/2014/FARS2014NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2015 <- read_csv("Raw Data/2015/FARS2015NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2016 <- read_csv("Raw Data/2016/FARS2016NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2017 <- read_csv("Raw Data/2017/FARS2017NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2018 <- read_csv("Raw Data/2018/FARS2018NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2019 <- read_csv("Raw Data/2019/FARS2019NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2020 <- read_csv("Raw Data/2020/FARS2020NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2021 <- read_csv("Raw Data/2021/FARS2021NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2022 <- read_csv("Raw Data/2022/FARS2022NationalAuxiliaryCSV/VEH_AUX.CSV", na = c(""))
veh2023 <- read_csv("Raw Data/2023/FARS2023NationalAuxiliaryCSV/FARS2023NationalAuxiliaryCSV/veh_aux.csv", na = c(""))
```

### NationalCSV

Importing the data files required significant manual effort considering that the naming conventions varied by certain time frames. Please review the following notes to understand the file extension variations:

| Time frame            | File Pathway Notes                                                                         |
|-------------------|-----------------------------------------------------|
| 2007-2008, 2023       | Contains extra subfolder                                                                   |
| 2007-2014             | All upper case                                                                             |
| 2015, 2018, 2020-2023 | All lower case                                                                             |
| 2016-2017             | Accident is in lower case and both Person and Vehicle are in proper spelling form.         |
| 2019                  | Accident is in lower case, Person is in proper spelling form, and Vehicle is in lower case |
| 2022-2023             | .csv file extension is in lowercase                                                        |

: NationalCSV Import Notes

```{r}
#| label: import-national-csv-data
#| message: false
#| warning: false
#| echo: true  


accident2007 <- read_csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/ACCIDENT.CSV")
accident2008 <- read_csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/ACCIDENT.CSV")
accident2009 <- read_csv("Raw Data/2009/FARS2009NationalCSV/ACCIDENT.CSV")
accident2010 <- read_csv("Raw Data/2010/FARS2010NationalCSV/ACCIDENT.CSV")
accident2011 <- read_csv("Raw Data/2011/FARS2011NationalCSV/ACCIDENT.CSV")
accident2012 <- read_csv("Raw Data/2012/FARS2012NationalCSV/ACCIDENT.CSV")
accident2013 <- read_csv("Raw Data/2013/FARS2013NationalCSV/ACCIDENT.CSV")
accident2014 <- read_csv("Raw Data/2014/FARS2014NationalCSV/ACCIDENT.CSV")
accident2015 <- read_csv("Raw Data/2015/FARS2015NationalCSV/accident.CSV")
accident2016 <- read_csv("Raw Data/2016/FARS2016NationalCSV/accident.CSV")
accident2017 <- read_csv("Raw Data/2017/FARS2017NationalCSV/accident.CSV")
accident2018 <- read_csv("Raw Data/2018/FARS2018NationalCSV/accident.CSV")
accident2019 <- read_csv("Raw Data/2019/FARS2019NationalCSV/accident.CSV")
accident2020 <- read_csv("Raw Data/2020/FARS2020NationalCSV/accident.CSV")
accident2021 <- read_csv("Raw Data/2021/FARS2021NationalCSV/accident.CSV")
accident2022 <- read_csv("Raw Data/2022/FARS2022NationalCSV/accident.csv")
accident2023 <- read_csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/accident.csv")
person2007 <- read_csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/PERSON.CSV")
person2008 <- read_csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/PERSON.CSV")
person2009 <- read_csv("Raw Data/2009/FARS2009NationalCSV/PERSON.CSV")
person2010 <- read_csv("Raw Data/2010/FARS2010NationalCSV/PERSON.CSV")
person2011 <- read_csv("Raw Data/2011/FARS2011NationalCSV/PERSON.CSV")
person2012 <- read_csv("Raw Data/2012/FARS2012NationalCSV/PERSON.CSV")
person2013 <- read_csv("Raw Data/2013/FARS2013NationalCSV/PERSON.CSV")
person2014 <- read_csv("Raw Data/2014/FARS2014NationalCSV/PERSON.CSV")
person2015 <- read_csv("Raw Data/2015/FARS2015NationalCSV/person.CSV")
person2016 <- read_csv("Raw Data/2016/FARS2016NationalCSV/Person.CSV")
person2017 <- read_csv("Raw Data/2017/FARS2017NationalCSV/Person.CSV")
person2018 <- read_csv("Raw Data/2018/FARS2018NationalCSV/person.CSV")
person2019 <- read_csv("Raw Data/2019/FARS2019NationalCSV/Person.CSV")
person2020 <- read_csv("Raw Data/2020/FARS2020NationalCSV/person.CSV")
person2021 <- read_csv("Raw Data/2021/FARS2021NationalCSV/person.CSV")
person2022 <- read_csv("Raw Data/2022/FARS2022NationalCSV/person.csv")
person2023 <- read_csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/person.csv")
vehicle2007 <- read_csv("Raw Data/2007/FARS2007NationalCSV/FARS2007NationalCSV/VEHICLE.CSV")
vehicle2008 <- read_csv("Raw Data/2008/FARS2008NationalCSV/FARS2008NationalCSV/VEHICLE.CSV")
vehicle2009 <- read_csv("Raw Data/2009/FARS2009NationalCSV/VEHICLE.CSV")
vehicle2010 <- read_csv("Raw Data/2010/FARS2010NationalCSV/VEHICLE.CSV")
vehicle2011 <- read_csv("Raw Data/2011/FARS2011NationalCSV/VEHICLE.CSV")
vehicle2012 <- read_csv("Raw Data/2012/FARS2012NationalCSV/VEHICLE.CSV")
vehicle2013 <- read_csv("Raw Data/2013/FARS2013NationalCSV/VEHICLE.CSV")
vehicle2014 <- read_csv("Raw Data/2014/FARS2014NationalCSV/VEHICLE.CSV")
vehicle2015 <- read_csv("Raw Data/2015/FARS2015NationalCSV/vehicle.CSV")
vehicle2016 <- read_csv("Raw Data/2016/FARS2016NationalCSV/Vehicle.CSV")
vehicle2017 <- read_csv("Raw Data/2017/FARS2017NationalCSV/Vehicle.CSV")
vehicle2018 <- read_csv("Raw Data/2018/FARS2018NationalCSV/vehicle.CSV")
vehicle2019 <- read_csv("Raw Data/2019/FARS2019NationalCSV/vehicle.CSV")
vehicle2020 <- read_csv("Raw Data/2020/FARS2020NationalCSV/vehicle.CSV")
vehicle2021 <- read_csv("Raw Data/2021/FARS2021NationalCSV/vehicle.CSV")
vehicle2022 <- read_csv("Raw Data/2022/FARS2022NationalCSV/vehicle.csv")
vehicle2023 <- read_csv("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/vehicle.csv")

```

# Preliminary Data Cleaning

This section refers to steps taken before data cleaning such as stacking AUX and National datasets, respectively, and selecting their respective years.

## AUX Stacked Dataframes

-   The ls() function searches for my environment for any object starting with "acc".
-   mget() takes the list of names I defined and gets the actual data

```{r}
#| label: acc-stacked
#| message: false
#| warning: false
#| echo: true  

# define acc data frames from environment
acc_list <- mget(ls(pattern = "^acc20"))
 
# bind data frames together
stacked_AUX_acc <- acc_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, FATALS, A_REGION, A_RU)



```

```{r}
#| label: per-stacked
#| message: false
#| warning: false
#| echo: true  

# define per data frames in environment
per_list <- mget(ls(pattern = "^per20"))

# bind data frames
stacked_AUX_per <- per_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, VEH_NO, PER_NO, A_PTYPE, A_RESTUSE, A_HELMUSE, A_PERINJ, A_AGE4, A_HRACE)

```

```{r}
#| label: veh-stacked
#| message: false
#| warning: false
#| echo: true  

# define veh data frames in environment
veh_list <- mget(ls(pattern = "^veh20"))

# bind data frames
stacked_AUX_veh <- veh_list %>%
  bind_rows(.id = "source_year") %>% 
  select(ST_CASE, YEAR, VEH_NO)

```

## National Stacked Data frames

Note to self: The accident files inherently contain the "year" column.

```{r}
#| label: accident-stacked
#| message: false
#| warning: false
#| echo: true  

# define accident data frames in environment and convert all columns to characters
accident_list <- mget(ls(pattern = "^accident[0-9]")) %>% 
  map(~ .x %>% mutate(across(everything(), as.character)))


# bind data frames
stacked_national_accident <- accident_list %>% 
  bind_rows(.id = "source_year") %>% 
  select(any_of(c(
    "ST_CASE", 
    "YEAR", 
    "STATE", 
    "COUNTY", 
    "VEH_NO", 
    "PER_NO", 
    "PERSONS", 
### Geographic Indicators
    "RUR_URB", 
    "RUR_URBNAME",
    "A_RU",
    "longitude", 
    "longitud", 
    "latitude", 
    "LATITUDE", 
    "LONGITUDE", 
    "LONGITUD", 
  "FUNC_SYS",
  "FUNC_SYS 1",
  "FUNC_SYS 2",
  "FUNC_SYS 3",
  "FUNC_SYS 4",
  "FUNC_SYS 5",
  "FUNC_SYS 6",
  "FUNC_SYS 7",
"ROAD_FNC",
### Time of Accident
  "MONTH",
  "MONTHNAME",
  "DAY",
  "DAYNAME",
  "DAY_WEEK",
  "DAY_WEEKNAME",
  "YEAR",
  "HOUR",
  "MINUTE",
### EMS variables
"NOT_HOUR",
"NOT_MIN",
"NOT_MINUTE",
"ARR_HOUR",
"ARR_MIN",
"HOSP_HR",
"HOSP_MN",
"FATALS"

  )))


```

Note to self: The person files inherently DO NOT contain the "year" column. And thus, I will need to manually create said column, and add to each year.

I used AI to help me create a function to add the year column

```{r}
#| label: person-years-function
#| message: false
#| warning: false
#| echo: true  

# Get the names of all objects in your environment that contain a year (4 digits)
person_names <- ls(pattern = "^person\\d{4}")

# Create a named list of these data frames
person_list <- mget(person_names)

# Add year column to each person data frame via function
person_list_updated <- lapply(names(person_list), function(df_name) {
  
  # Extract the year from the string (e.g., "2020")
  year_val <- str_extract(df_name, "\\d{4}")
  
  # Access the data frame
  df <- person_list[[df_name]]
  
  # Add the column
  df$YEAR <- year_val
  
  return(df)
})

# 4. Rename the list elements to match original names
names(person_list_updated) <- person_names

# update R environment with 
list2env(person_list_updated, envir = .GlobalEnv)
```

```{r}
#| label: person-stacked
#| message: false
#| warning: false
#| echo: true  

# define person data frames in environment and convert all columns to characters
person_list <- mget(ls(pattern = "^person[0-9]")) %>% 
  map(~ .x %>% mutate(across(everything(), as.character)))

# bind data frames
stacked_national_person <- person_list %>% 
  bind_rows(.id = "source_year") %>% 
  select(any_of(c(
    "ST_CASE", 
    "YEAR", 
    "STATE", 
    "COUNTY", 
    "VEH_NO", 
    "PER_NO", 
    "RUR_URB", 
    "RUR_URBNAME", 
    "PERSONS", 
    "FATALS", 
    "A_RU", 
    "longitude", 
    "longitud", 
    "latitude", 
    "LATITUDE", 
    "LONGITUDE", 
    "LONGITUD",
### Demographics
    "AGE",
    "SEX",
### behavioral variables
    "ALC_STATUS",
    "ALC_RES",
    "DRUGS",
    "DRUGSNAME",
    "DSTATUS",
    "REST_USE",
    "REST_MIS",
    "REST_MISNAME",
    "REST_USENAME",
    "HELM_USE",
    "HELM_USENAME",
### EMS variables
"INJ_SEV",
"INJ_SEVNAME",
"HOSPITAL",
"HOSPITALNAME",
"DOA",
"DOANAME",
"DEATH_DA",
"DEATH_MN",
"DEATH_YR",
"LAG_HRS",
"LAG_MINS",
"BODY_TYP",
"BODY_TYPNAME"
  )))


```

Note to self: The vehicle files DO NOT contain the "year" column. And thus, I will need to manually create said column, and add to each year.

I modified the previous person code chunk for vehicle.

```{r}
#| label: vehicle-years-function
#| message: false
#| warning: false
#| echo: true  

# Get the names of all objects in your environment that contain a year (4 digits)
vehicle_names <- ls(pattern = "^vehicle\\d{4}")

# Create a named list of these data frames
vehicle_list <- mget(vehicle_names)

# Add year column to each vehicle data frame via function
vehicle_list_updated <- lapply(names(vehicle_list), function(df_name) {
  
  # Extract the year from the string (e.g., "2020")
  year_val <- str_extract(df_name, "\\d{4}")
  
  # Access the data frame
  df <- vehicle_list[[df_name]]
  
  # Add the column
  df$YEAR <- year_val
  
  return(df)
})

# 4. Rename the list elements to match original names
names(vehicle_list_updated) <- vehicle_names

# update R environment with 
list2env(vehicle_list_updated, envir = .GlobalEnv)
```

```{r}
#| label: vehicle-stacked
#| message: false
#| warning: false
#| echo: true  

# define vehicle data frames in environment and convert all columns to characters
vehicle_list <- mget(ls(pattern = "^vehicle[0-9]")) %>% 
  map(~ .x %>% mutate(across(everything(), as.character)))

# bind data frames
stacked_national_vehicle <- vehicle_list %>% 
  bind_rows(.id = "source_year") %>% 
  select(any_of(c(
    "ST_CASE", 
    "YEAR", 
    "STATE", 
    "COUNTY", 
    "VEH_NO", 
    "PER_NO", 
    "RUR_URB", 
    "RUR_URBNAME", 
    "PERSONS", 
    "FATALS", 
    "A_RU",
    "longitude", 
    "longitud", 
    "latitude", 
    "LATITUDE", 
    "LONGITUDE", 
    "LONGITUD",
### behavioral variables
    "DR_DRINK",
    "SPEEDREL",
    "PREV_SPD",
"DR_ZIP",
"SPEC_USE"
  )))

```

## Remove dataframes

I remove obsolete dataframes from my environment to clear up memory and to also remove unneeded dataframes for data exporting purposes.

```{r}
#| label: remove-dataframes-old
#| message: false
#| warning: false
#| echo: true  

# remove obsolete AUX data
rm(list = ls(pattern = "^acc20")) 
rm(list = ls(pattern = "^per20"))
rm(list = ls(pattern = "^veh20"))

# remove obsolete National data
rm(list = ls(pattern = "accident20"))
rm(list = ls(pattern = "person20"))
rm(list = ls(pattern = "vehicle20"))
rm(list = ls(pattern = "_list"))
rm(list = c("person_names","vehicle_names"))


```

# Data Cleaning

| Stacked Dataset              | Verify Correct Logic? | Notes                                                  |
|-------------------|-------------------|-----------------------------------|
| stacked_national_person_v2   | ✅                    | Not sure what logic to put for LAG_HRS due to the note |
| PER_AUX                      | ✅                    |                                                        |
| stacked_national_accident_v2 | ✅                    |                                                        |
| stacked_national_vehicle_v2  | ✅                    |                                                        |

## National person data cleaning

```{r}
#| label: national-person-data-cleaning
#| message: false
#| warning: false
#| echo: true  

stacked_national_person_v2 <- stacked_national_person %>% 
  mutate(RUR_URBNAME = if_else(RUR_URB >= 8 & RUR_URB <= 9, "", RUR_URBNAME),
         RUR_URB = if_else(RUR_URB >= 8 & RUR_URB <= 9, "", RUR_URB),
         REST_USE = if_else(REST_USE >= 96 & REST_USE <= 99, "", REST_USE),
         REST_USENAME = if_else(REST_USE >= 96 & REST_USE <= 99, "", REST_USENAME),
         REST_MIS = if_else(REST_MIS >= 8 & REST_MIS <= 9, "", REST_MIS),
         REST_MISNAME = if_else(REST_MISNAME >= 8 & REST_MISNAME <= 9, "", REST_MISNAME),
         HELM_USE = if_else(HELM_USE >= 96 & HELM_USE <= 99, "", HELM_USE),
         HELM_USENAME = if_else(HELM_USENAME >= 96 & HELM_USENAME <= 99, "", HELM_USENAME),
         BODY_TYP = if_else(BODY_TYP >= 98 & BODY_TYP <= 99, "", BODY_TYP),
         BODY_TYPNAME = if_else(BODY_TYPNAME >= 98 & BODY_TYPNAME <= 99, "", BODY_TYPNAME),
         INJ_SEV = if_else(INJ_SEV >= 8 & INJ_SEV <= 9, "", INJ_SEV),
         INJ_SEVNAME = if_else(INJ_SEVNAME >= 8 & INJ_SEVNAME <= 9, "", INJ_SEVNAME),
         SEX = if_else(SEX >= 8 & SEX <= 9, "", SEX),
         AGE = if_else(AGE >= 115, "", AGE),
         ALC_STATUS = if_else(ALC_STATUS >= 8 & ALC_STATUS <= 9, "", ALC_STATUS),
         ALC_RES = if_else(ALC_RES > 94 & YEAR != 2015 | 
                           ALC_RES > 940 & YEAR == 2015, "", ALC_RES),
         DRUGS = if_else(DRUGS >= 8 & DRUGS <= 9, "", DRUGS),
         DSTATUS = if_else(DSTATUS >= 8 & DSTATUS <= 9, "", DSTATUS),
         DOA = if_else(DOA == "9", "", DOA),
         DEATH_MN = if_else(DEATH_MN > 59, "", DEATH_MN),
         DEATH_DA = if_else(DEATH_DA > 31, "", DEATH_DA),
         DEATH_YR = if_else(DEATH_YR > 2050, "", DEATH_YR),
         LAG_HRS = if_else(LAG_HRS > 719, "", LAG_HRS), ## note says: "/*Different before 2010*/". What should I put for logic?    
         LAG_MINS = if_else(LAG_MINS > 59, "", LAG_MINS)
)

```

## AUX person files data cleaning

```{r}
#| label: AUX-person-data-cleaning
#| message: false
#| warning: false
#| echo: true  

stacked_AUX_per_v2 <- stacked_AUX_per %>% 
  mutate(A_HRACE = if_else(A_HRACE == 9, "", as.character(A_HRACE)))

```

## National Accident files

I combined Trish data cleaning 2014 and 2015 as noted below.

```{r}
#| label: national-accident-data-cleaning
#| message: false
#| warning: false
#| echo: true  

stacked_national_accident_v2 <- stacked_national_accident %>% 
  mutate(COUNTY = if_else(COUNTY > 996, "", COUNTY),
         LATITUDE = if_else(LATITUDE >= 77.0, "", LATITUDE),
         LONGITUD = if_else(LONGITUD > 700, "", LONGITUD),
         ROAD_FNC = if_else(
           ROAD_FNC == 9 & YEAR <= 2014 | 
           ROAD_FNC == 19 & YEAR <= 2014 | 
           ROAD_FNC == 99 & YEAR <= 2014, ## 2014 and before
           "", ROAD_FNC),
         FUNC_SYS = if_else(FUNC_SYS >= 96 & YEAR >= 2015, "", FUNC_SYS), ## 2015 and after
         DAY = if_else(DAY > 31, "", DAY),
         DAY_WEEK = if_else(DAY_WEEK > 7, "", DAY_WEEK),
         MONTH = if_else(MONTH >12, "", MONTH),
         HOUR = if_else(HOUR >24, "", HOUR),
         MINUTE = if_else(MINUTE > 59, "", MINUTE),
         NOT_HOUR = if_else(NOT_HOUR > 24, "", NOT_HOUR),
         NOT_MIN = if_else(NOT_MIN > 59, "", NOT_MIN),
         ARR_HOUR = if_else(ARR_HOUR > 24, "", ARR_HOUR),
         ARR_MIN = if_else(ARR_MIN > 59, "", ARR_MIN),
         HOSP_HR = if_else(HOSP_HR > 24, "", HOSP_HR),
         HOSP_MN = if_else(HOSP_MN > 59, "", HOSP_MN)
  )



```

## National Vehicle files

```{r}
#| label: label_name
#| message: false
#| warning: false
#| echo: true  

stacked_national_vehicle_v2 <- stacked_national_vehicle %>% 
  mutate(SPEEDREL = if_else( SPEEDREL == 9, "", SPEEDREL),
         PREV_SPD = if_else(PREV_SPD > 97, "", PREV_SPD),
         SPEC_USE = if_else(SPEC_USE>=98, "", SPEC_USE),
         DR_ZIP = if_else(DR_ZIP >= 99997, "", DR_ZIP)
  )

```

### Remove old stacked dataframes from environment

```{r}
#| label: remove-old-stacked-dataframes
#| message: false
#| warning: false
#| echo: true  

rm(stacked_national_accident)
rm(stacked_national_person)
rm(stacked_national_vehicle)
rm(stacked_AUX_per)
```

# Merging Stacked Datasets

## Convert all variables to character

```{r}
#| label: convert-variables-character
#| message: false
#| warning: false
#| echo: true  

# 1. Identify all objects in the global environment that are data frames
df_names <- names(which(sapply(ls(), function(x) is.data.frame(get(x)))))

# 2. Loop through each data frame, convert columns, and save them back
for (name in df_names) {
  # Get the actual data frame object
  df <- get(name)
  
  # Convert all columns to character (using df[] preserves the data frame structure)
  df[] <- lapply(df, as.character)
  
  # Assign the modified data frame back to its original name in the global environment
  assign(name, df, envir = .GlobalEnv)
}
```

### remove values

```{r}
#| label: remove-values
#| message: false
#| warning: false
#| echo: true  

rm(df_names)
rm(name)
```

## Merge datasets

In this step, I merge the following datasets together using the stacked national person file as the base.

-   AUX_person
-   AUX_accident
-   national_accident

```{r}
#| label: merge-datasets
#| message: false
#| warning: false
#| echo: true  


df <- stacked_national_person_v2 %>% 
  left_join(stacked_national_accident_v2, by = c("ST_CASE", "YEAR")) %>% 
  left_join(stacked_AUX_per_v2, by = c("ST_CASE","YEAR","VEH_NO","PER_NO")) %>% 
  left_join(stacked_AUX_acc, by = c("ST_CASE","YEAR")) %>% 
  select(-ends_with(".x"), -ends_with(".y"))  ## remove duplicate columns


```

## Analytical Dataset restrictions
df is the completely merged dataset. I restrict the dataset to fatal cases in private vehicle uses and occupants inside cars.

```{r}
#| label: df-dataset-restrictions
#| message: false
#| warning: false
#| echo: true  

df <- df %>% 
  filter(INJ_SEV == 4) %>%  ## dead people only
  filter(YEAR >= 2010) %>%  ## filter to years 2010+
## vehicle filters
  filter(BODY_TYP %in% c(1:8) |
          BODY_TYP == 17 | ## passenger cars
         BODY_TYP %in% c(14:16) | ## SUVs
           BODY_TYP == 19 |
         BODY_TYP == 20 | ## minivans
         BODY_TYP %in% c(30:39)) %>%  ## pickups
## driver or occupant in car
  filter(A_PTYPE %in% c(1:2)) ##%>% 
## private vehicle use [NOTE: CANNOT APPLY FILTER YET BECAUSE WE DID NOT MERGE IN NATIONAL VEHICLE FILE. WAITING FOR INSTRUCTIONS FROM TRISH]
 ## filter(SPEC_USE == 0)


```


# Data Export

I export all the datasets so I can upload them to BOX.

## df split
I split the dataset into three smaller subdatasets based on the following year ranges:

* 2010-2015
* 2016-2020
* 2021-2024

This is so we can verify the data in Excel.
```{r}
#| label: df-split
#| message: false
#| warning: false
#| echo: true  

df_2010_2015 <- df %>% 
  filter(YEAR %in% c(2010:2015))

df_2016_2020 <- df %>% 
  filter(YEAR %in% c(2016:2020))

df_2021_2024 <- df %>% 
  filter(YEAR %in% c(2021:2024))
```

## Data Export
```{r}
#| label: data_export
#| message: false
#| warning: false
#| echo: true

# I filter the environment for all data frame objects and create a list
all_data_frames <- mget(ls()) %>% 
  keep(is.data.frame)


# I walk through the entire list and tell R to write .csv files into the "Export" folder.
iwalk(all_data_frames, ~ write_csv(.x, paste0("Exports/NHTSA Data Cleaning/", .y, ".csv"), na =""))



```

