---
title: "NHTSA Analysis"
author: "Patrick Torralba"
format: 
  html:
    theme: flatly
    fontsize: 0.8em
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 4
    smooth-scroll: TRUE
    embed-resources: TRUE
    fig-align: center
editor: visual
---

# Set-up

## Library

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true 

library(tidyverse)
library(kableExtra)
library(sf)
library(rnaturalearth)
library(scales) ## add commas 
library(gtsummary) ## needed for tbl_regression

```

## Load Data

```{r}
#| label: load-data
#| message: false
#| warning: false
#| echo: true  


df_v2 <- read.csv("Exports/df_v2.csv")

```

## Data Cleaning

We want to remove NA geographic values and change "location" to a factor. "location" has urban set as the reference level.

```{r}
#| label: data-clean
#| message: false
#| warning: false
#| echo: true  

df_v2 <- df_v2 %>% 
  filter(!is.na(location)) %>% 
  mutate(location = fct_relevel(as.factor(location),"Urban"))

```

# Data Analysis & Visualization

## Rural vs Urban Fatality Aggregation

```{r}
#| label: rur_urb_agg
#| message: false
#| warning: false
#| echo: true  


rur_urb_agg <- df_v2 %>% 
  group_by(YEAR, location) %>% 
  summarize(
    total_persons = sum(PER_NO),
    total_fatals = sum(FATALS.x)) 

```

## Rural vs Urban Fatality Time-series

```{r}
#| label: fig-time-series
#| fig-cap: "Rural vs Urban Fatality Time-series"
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: true

ggplot(rur_urb_agg, aes(x = YEAR, y = total_fatals)) +
  geom_col() +
  geom_line(aes(color = location, group = location)) +
scale_x_continuous(breaks = seq(min(rur_urb_agg$YEAR), max(rur_urb_agg$YEAR), by = 1)) +  
  labs(
    x = "Year",
    y = "Total Fatalalities",
    color = "Location"
  ) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


```

## Rural vs Urban Fatality Table

```{r}
#| label: tbl-fatality
#| tbl-cap: "Rural vs Urban Fatalities"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true


kable(rur_urb_agg,
      format = "html",
      col.names = c("Year","Location","Total Persons", "Total Fatals")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))


```

# Poisson Regression Modeling

We use a poisson regression because we want to calculate the rate probability over a fixed amount of time (e.g fatality count over 1 year).

```{r}
#| label: poisson-regression
#| message: false
#| warning: false
#| echo: true  

fatality_model <- glm(total_fatals ~ location,
                      data = rur_urb_agg,
                      family = poisson(link = "log")
  
)

fatality_model
```

```{r}
#| label: tbl-fatality-model-exp
#| tbl-cap: "Unadjusted Fatality Model [Exponent Form]"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

fatality_model %>% 
  tbl_regression(
    intercept = TRUE,
    exponentiate = TRUE)
```
