---
title: "NHTSA Analysis"
author: "Patrick Torralba"
format: 
  html:
    theme: flatly
    fontsize: 0.8em
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 4
    smooth-scroll: TRUE
    embed-resources: TRUE
    fig-align: center
editor: visual
---

# Set-up

## Library

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true 

library(tidyverse)
library(kableExtra)
library(sf)
library(rnaturalearth)
library(scales) ## add commas 
library(gtsummary) ## needed for tbl_regression

```

## Load Data

```{r}
#| label: load-data
#| message: false
#| warning: false
#| echo: true  


df_v2 <- read.csv("Exports/df_v2.csv")

```

## Data Cleaning

We want to remove NA geographic values and change "location" to a factor. "location" has urban set as the reference level.

```{r}
#| label: data-clean
#| message: false
#| warning: false
#| echo: true  

df_v2 <- df_v2 %>% 
  filter(!is.na(location)) %>% 
  mutate(location = fct_relevel(as.factor(location),"Urban"))

```

# Data Analysis & Visualization

## Rural vs Urban Fatality Aggregation

```{r}
#| label: rur_urb_agg
#| message: false
#| warning: false
#| echo: true  


rur_urb_agg <- df_v2 %>% 
  group_by(YEAR, location) %>% 
  summarize(
    total_persons = sum(PER_NO),
    total_fatals = sum(FATALS.x)) 

```

## Rural vs Urban Fatality Time-series

```{r}
#| label: fig-time-series
#| fig-cap: "Rural vs Urban Fatality Time-series"
#| fig-cap-location: top
#| message: false
#| warning: false
#| echo: true

ggplot(rur_urb_agg, aes(x = YEAR, y = total_fatals)) +
  geom_col() +
  geom_line(aes(color = location, group = location)) +
scale_x_continuous(breaks = seq(min(rur_urb_agg$YEAR), max(rur_urb_agg$YEAR), by = 1)) +  
  labs(
    x = "Year",
    y = "Total Fatalalities",
    color = "Location"
  ) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))


```

## Rural vs Urban Fatality Table

```{r}
#| label: tbl-fatality
#| tbl-cap: "Rural vs Urban Fatalities"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true


kable(rur_urb_agg,
      format = "html",
      col.names = c("Year","Location","Total Persons", "Total Fatals")) %>% 
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))


```

# Poisson Regression Modeling

We use a poisson regression because we want to calculate the rate probability over a fixed amount of time (e.g fatality count over 1 year).

## Unadjusted Model
```{r}
#| label: poisson-regression-unadjusted
#| message: false
#| warning: false
#| echo: true  

model1 <- glm(total_fatals ~ location,
                 data = rur_urb_agg, 
                 family = poisson(link = "log"))

model1
```

The unadjusted model only accounts for the total fatalities affected by location.


```{r}
#| label: tbl-fatality-model-exp
#| tbl-cap: "Unadjusted Fatality Model [Exponent Form]"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

fatality_model %>% 
  tbl_regression(
    intercept = TRUE,
    exponentiate = TRUE)
```

## Quasi-Poisson Regression Model for Overdispersion

Before proceeding further with these results, we must first test for overdispersion by reviewing the ratio of the residual deviance to its degree of freedoms. If the model ratio is extremely greater than 1 then that is a sign of overdispersion.

```{r}
#| label: overdispersion-test
#| message: false
#| warning: false
#| echo: true  

model_ratio <- (model1$deviance / model1$df.residual)

model_ratio

model_ratio > 10

```

Since we verified that this model is overdispersed, we now must deal with it using a quasi-poisson regression model

```{r}
#| label: poisson-regression-model
#| message: false
#| warning: false
#| echo: true  

model1.qp <- glm(total_fatals ~ location,
                 data = rur_urb_agg, 
                 family = quasipoisson(link = "log"))

model1.qp
```

It appears that are unadjusted model displays there is no statistically significant association between fatality outcome and rural location.
```{r}
#| label: tbl-poisson-regression-model-qp
#| tbl-cap: "Quasi-poisson Regression Model"
#| tbl-cap-location: top
#| message: false
#| warning: false
#| echo: true

model1.qp %>% 
  tbl_regression(
    exponentiate = TRUE,
    label = (location ~ "Location")
  )

```

