---
title: "NHTSA Data Cleaning"
author: "Patrick Torralba"
format:
  html:
    theme: flatly
    fontsize: 0.8em
    code-fold: FALSE
    code-tools: TRUE
    code-copy: TRUE
    toc: TRUE
    toc-depth: 4
    smooth-scroll: TRUE
    embed-resources: TRUE
    fig-align: center
editor: visual
---

# Set-up

## Library

```{r}
#| label: library
#| message: false
#| warning: false
#| echo: true  

library(tidyverse)
library(scales)
library(data.table)
```

## Import Data

I used Excel to help create functions to mass import the .csv files for both AUX and National CSV data.

### National Auxillary CSV Data

Please note that the file path for 2023 is slightly different in that the extension is lower case and has an additional folder.

```{r}
#| label: import-AUX-data
#| message: false
#| warning: false
#| echo: true  
#| cache: true

acc2007 <- fread("Raw Data/2007/FARS2007NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2008 <- fread("Raw Data/2008/FARS2008NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2009 <- fread("Raw Data/2009/FARS2009NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2010 <- fread("Raw Data/2010/FARS2010NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2011 <- fread("Raw Data/2011/FARS2011NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2012 <- fread("Raw Data/2012/FARS2012NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2013 <- fread("Raw Data/2013/FARS2013NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2014 <- fread("Raw Data/2014/FARS2014NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2015 <- fread("Raw Data/2015/FARS2015NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2016 <- fread("Raw Data/2016/FARS2016NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2017 <- fread("Raw Data/2017/FARS2017NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2018 <- fread("Raw Data/2018/FARS2018NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2019 <- fread("Raw Data/2019/FARS2019NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2020 <- fread("Raw Data/2020/FARS2020NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2021 <- fread("Raw Data/2021/FARS2021NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2022 <- fread("Raw Data/2022/FARS2022NationalAuxiliaryCSV/ACC_AUX.CSV", na.strings = "")
acc2023 <- fread("Raw Data/2023/FARS2023NationalAuxiliaryCSV/acc_aux.csv", na.strings = "")
per2007 <- fread("Raw Data/2007/FARS2007NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2008 <- fread("Raw Data/2008/FARS2008NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2009 <- fread("Raw Data/2009/FARS2009NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2010 <- fread("Raw Data/2010/FARS2010NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2011 <- fread("Raw Data/2011/FARS2011NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2012 <- fread("Raw Data/2012/FARS2012NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2013 <- fread("Raw Data/2013/FARS2013NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2014 <- fread("Raw Data/2014/FARS2014NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2015 <- fread("Raw Data/2015/FARS2015NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2016 <- fread("Raw Data/2016/FARS2016NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2017 <- fread("Raw Data/2017/FARS2017NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2018 <- fread("Raw Data/2018/FARS2018NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2019 <- fread("Raw Data/2019/FARS2019NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2020 <- fread("Raw Data/2020/FARS2020NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2021 <- fread("Raw Data/2021/FARS2021NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2022 <- fread("Raw Data/2022/FARS2022NationalAuxiliaryCSV/PER_AUX.CSV", na.strings = "")
per2023 <- fread("Raw Data/2023/FARS2023NationalAuxiliaryCSV/per_aux.csv", na.strings = "")
veh2007 <- fread("Raw Data/2007/FARS2007NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2008 <- fread("Raw Data/2008/FARS2008NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2009 <- fread("Raw Data/2009/FARS2009NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2010 <- fread("Raw Data/2010/FARS2010NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2011 <- fread("Raw Data/2011/FARS2011NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2012 <- fread("Raw Data/2012/FARS2012NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2013 <- fread("Raw Data/2013/FARS2013NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2014 <- fread("Raw Data/2014/FARS2014NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2015 <- fread("Raw Data/2015/FARS2015NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2016 <- fread("Raw Data/2016/FARS2016NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2017 <- fread("Raw Data/2017/FARS2017NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2018 <- fread("Raw Data/2018/FARS2018NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2019 <- fread("Raw Data/2019/FARS2019NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2020 <- fread("Raw Data/2020/FARS2020NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2021 <- fread("Raw Data/2021/FARS2021NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2022 <- fread("Raw Data/2022/FARS2022NationalAuxiliaryCSV/VEH_AUX.CSV", na.strings = "")
veh2023 <- fread("Raw Data/2023/FARS2023NationalAuxiliaryCSV/veh_aux.csv", na.strings = "")
```

### NationalCSV

Importing the data files required significant manual effort considering that the naming conventions varied by certain time frames. Please review the following notes to understand the file extension variations:

| Time frame | File Pathway Notes |
|-------------------------|-----------------------------------------------|
| 2007-2008, 2023 | Contains extra subfolder |
| 2007-2014 | All upper case |
| 2015, 2018, 2020-2023 | All lower case |
| 2016-2017 | Accident is in lower case and both Person and Vehicle are in proper spelling form. |
| 2019 | Accident is in lower case, Person is in proper spelling form, and Vehicle is in lower case |
| 2022-2023 | .csv file extension is in lowercase |

: NationalCSV Import Notes

```{r}
#| label: import-national-csv-data
#| message: false
#| warning: false
#| echo: true  
#| cache: true


accident2007 <- fread("Raw Data/2007/FARS2007NationalCSV/ACCIDENT.CSV")
accident2008 <- fread("Raw Data/2008/FARS2008NationalCSV/ACCIDENT.CSV")
accident2009 <- fread("Raw Data/2009/FARS2009NationalCSV/ACCIDENT.CSV")
accident2010 <- fread("Raw Data/2010/FARS2010NationalCSV/ACCIDENT.CSV")
accident2011 <- fread("Raw Data/2011/FARS2011NationalCSV/ACCIDENT.CSV")
accident2012 <- fread("Raw Data/2012/FARS2012NationalCSV/ACCIDENT.CSV")
accident2013 <- fread("Raw Data/2013/FARS2013NationalCSV/ACCIDENT.CSV")
accident2014 <- fread("Raw Data/2014/FARS2014NationalCSV/ACCIDENT.CSV")
accident2015 <- fread("Raw Data/2015/FARS2015NationalCSV/accident.CSV")
accident2016 <- fread("Raw Data/2016/FARS2016NationalCSV/accident.CSV")
accident2017 <- fread("Raw Data/2017/FARS2017NationalCSV/accident.CSV")
accident2018 <- fread("Raw Data/2018/FARS2018NationalCSV/accident.CSV")
accident2019 <- fread("Raw Data/2019/FARS2019NationalCSV/accident.CSV")
accident2020 <- fread("Raw Data/2020/FARS2020NationalCSV/accident.CSV")
accident2021 <- fread("Raw Data/2021/FARS2021NationalCSV/accident.CSV")
accident2022 <- fread("Raw Data/2022/FARS2022NationalCSV/accident.csv")
accident2023 <- fread("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/accident.csv")
person2007 <- fread("Raw Data/2007/FARS2007NationalCSV/PERSON.CSV")
person2008 <- fread("Raw Data/2008/FARS2008NationalCSV/PERSON.CSV")
person2009 <- fread("Raw Data/2009/FARS2009NationalCSV/PERSON.CSV")
person2010 <- fread("Raw Data/2010/FARS2010NationalCSV/PERSON.CSV")
person2011 <- fread("Raw Data/2011/FARS2011NationalCSV/PERSON.CSV")
person2012 <- fread("Raw Data/2012/FARS2012NationalCSV/PERSON.CSV")
person2013 <- fread("Raw Data/2013/FARS2013NationalCSV/PERSON.CSV")
person2014 <- fread("Raw Data/2014/FARS2014NationalCSV/PERSON.CSV")
person2015 <- fread("Raw Data/2015/FARS2015NationalCSV/person.CSV")
person2016 <- fread("Raw Data/2016/FARS2016NationalCSV/Person.CSV")
person2017 <- fread("Raw Data/2017/FARS2017NationalCSV/Person.CSV")
person2018 <- fread("Raw Data/2018/FARS2018NationalCSV/person.CSV")
person2019 <- fread("Raw Data/2019/FARS2019NationalCSV/Person.CSV")
person2020 <- fread("Raw Data/2020/FARS2020NationalCSV/person.CSV")
person2021 <- fread("Raw Data/2021/FARS2021NationalCSV/person.CSV")
person2022 <- fread("Raw Data/2022/FARS2022NationalCSV/person.csv")
person2023 <- fread("Raw Data/2023/FARS2023NationalCSV/FARS2023NationalCSV/person.csv")
vehicle2007 <- fread("Raw Data/2007/FARS2007NationalCSV/VEHICLE.CSV")
vehicle2008 <- fread("Raw Data/2008/FARS2008NationalCSV/VEHICLE.CSV")
vehicle2009 <- fread("Raw Data/2009/FARS2009NationalCSV/VEHICLE.CSV")
vehicle2010 <- fread("Raw Data/2010/FARS2010NationalCSV/VEHICLE.CSV")
vehicle2011 <- fread("Raw Data/2011/FARS2011NationalCSV/VEHICLE.CSV")
vehicle2012 <- fread("Raw Data/2012/FARS2012NationalCSV/VEHICLE.CSV")
vehicle2013 <- fread("Raw Data/2013/FARS2013NationalCSV/VEHICLE.CSV")
vehicle2014 <- fread("Raw Data/2014/FARS2014NationalCSV/VEHICLE.CSV")
vehicle2015 <- fread("Raw Data/2015/FARS2015NationalCSV/vehicle.CSV")
vehicle2016 <- fread("Raw Data/2016/FARS2016NationalCSV/Vehicle.CSV")
vehicle2017 <- fread("Raw Data/2017/FARS2017NationalCSV/Vehicle.CSV")
vehicle2018 <- fread("Raw Data/2018/FARS2018NationalCSV/vehicle.CSV")
vehicle2019 <- fread("Raw Data/2019/FARS2019NationalCSV/vehicle.CSV")
vehicle2020 <- fread("Raw Data/2020/FARS2020NationalCSV/vehicle.CSV")
vehicle2021 <- fread("Raw Data/2021/FARS2021NationalCSV/vehicle.CSV")
vehicle2022 <- fread("Raw Data/2022/FARS2022NationalCSV/vehicle.csv")
vehicle2023 <- fread("Raw Data/2023/FARS2023NationalCSV/vehicle.csv")

```

# Preliminary Data Cleaning

This section refers to steps taken before data cleaning such as stacking AUX and National datasets, respectively, and selecting their respective years.

## AUX Stacked Dataframes

- The ls() function searches for my environment for any object starting with "acc".
- mget() takes the list of names I defined and gets the actual data

```{r}
#| label: acc-stacked
#| message: false
#| warning: false
#| echo: true  

# define acc data frames from environment
acc_list <- mget(ls(pattern = "^acc20"))
 
# bind data frames together
stacked_AUX_acc <- acc_list |> 
  bind_rows(.id = "source_year") |>  
  select(ST_CASE, YEAR, FATALS, A_REGION, A_RU)



```

```{r}
#| label: per-stacked
#| message: false
#| warning: false
#| echo: true  

# define per data frames in environment
per_list <- mget(ls(pattern = "^per20"))

# bind data frames
stacked_AUX_per <- per_list |> 
  bind_rows(.id = "source_year") |>  
  select(ST_CASE, YEAR, VEH_NO, PER_NO, A_PTYPE, A_RESTUSE, A_HELMUSE, A_PERINJ, A_AGE4, A_HRACE)

```

```{r}
#| label: veh-stacked
#| message: false
#| warning: false
#| echo: true  

# define veh data frames in environment
veh_list <- mget(ls(pattern = "^veh20"))

# bind data frames
stacked_AUX_veh <- veh_list |> 
  bind_rows(.id = "source_year") |>  
  select(ST_CASE, YEAR, VEH_NO)

```

## National Stacked Data frames

Note to self: The accident files inherently contain the "year" column.

```{r}
#| label: accident-stacked
#| message: false
#| warning: false
#| echo: true  
#| cache: true

# define accident data frames in environment and convert all columns to characters
accident_list <- mget(ls(pattern = "^accident[0-9]")) |>  
  map(~ .x |>  mutate(across(everything(), as.character)))


# bind data frames
stacked_national_accident <- accident_list |>  
  bind_rows(.id = "source_year") |>  
  select(any_of(c(
    "ST_CASE", 
    "YEAR", 
    "STATE", 
    "COUNTY", 
    "VEH_NO", 
    "PER_NO", 
    "PERSONS", 
### Geographic Indicators
    "RUR_URB", 
    "RUR_URBNAME",
    "A_RU",
    "longitude", 
    "longitud", 
    "latitude", 
    "LATITUDE", 
    "LONGITUDE", 
    "LONGITUD", 
  "FUNC_SYS",
  "FUNC_SYS 1",
  "FUNC_SYS 2",
  "FUNC_SYS 3",
  "FUNC_SYS 4",
  "FUNC_SYS 5",
  "FUNC_SYS 6",
  "FUNC_SYS 7",
"ROAD_FNC",
### Time of Accident
  "MONTH",
  "MONTHNAME",
  "DAY",
  "DAYNAME",
  "DAY_WEEK",
  "DAY_WEEKNAME",
  "YEAR",
  "HOUR",
  "MINUTE",
### EMS variables
"NOT_HOUR",
"NOT_MIN",
"NOT_MINUTE",
"ARR_HOUR",
"ARR_MIN",
"HOSP_HR",
"HOSP_MN",
"FATALS"

  )))


```

Note to self: The person files inherently DO NOT contain the "year" column. And thus, I will need to manually create said column, and add to each year.

I used AI to help me create a function to add the year column

```{r}
#| label: person-years-function
#| message: false
#| warning: false
#| echo: true  
#| cache: true

# Get the names of all objects in your environment that contain a year (4 digits)
person_names <- ls(pattern = "^person\\d{4}")

# Create a named list of these data frames
person_list <- mget(person_names)

# Add year column to each person data frame via function
person_list_updated <- lapply(names(person_list), function(df_name) {
  
  # Extract the year from the string (e.g., "2020")
  year_val <- str_extract(df_name, "\\d{4}")
  
  # Access the data frame
  df <- person_list[[df_name]]
  
  # Add the column
  df$YEAR <- year_val
  
  return(df)
})

# 4. Rename the list elements to match original names
names(person_list_updated) <- person_names

# update R environment with 
list2env(person_list_updated, envir = .GlobalEnv)
```

```{r}
#| label: person-stacked
#| message: false
#| warning: false
#| echo: true  
#| cache: true

# define person data frames in environment and convert all columns to characters
person_list <- mget(ls(pattern = "^person[0-9]")) |>  
  map(~ .x |>  mutate(across(everything(), as.character)))

# bind data frames
stacked_national_person <- person_list |>  
  bind_rows(.id = "source_year") |>  
  select(any_of(c(
    "ST_CASE", 
    "YEAR", 
    "STATE", 
    "COUNTY", 
    "VEH_NO", 
    "PER_NO", 
    "RUR_URB", 
    "RUR_URBNAME", 
    "PERSONS", 
    "FATALS", 
    "A_RU", 
    "longitude", 
    "longitud", 
    "latitude", 
    "LATITUDE", 
    "LONGITUDE", 
    "LONGITUD",
### Demographics
    "AGE",
    "SEX",
### behavioral variables
    "ALC_STATUS",
    "ALC_RES",
    "DRUGS",
    "DRUGSNAME",
    "DSTATUS",
    "REST_USE",
    "REST_MIS",
    "REST_MISNAME",
    "REST_USENAME",
    "HELM_USE",
    "HELM_USENAME",
### EMS variables
"INJ_SEV",
"INJ_SEVNAME",
"HOSPITAL",
"HOSPITALNAME",
"DOA",
"DOANAME",
"DEATH_DA",
"DEATH_MN",
"DEATH_YR",
"LAG_HRS",
"LAG_MINS",
"BODY_TYP",
"BODY_TYPNAME"
  )))


```

Note to self: The vehicle files DO NOT contain the "year" column. And thus, I will need to manually create said column, and add to each year.

I modified the previous person code chunk for vehicle.

```{r}
#| label: vehicle-years-function
#| message: false
#| warning: false
#| echo: true  

# Get the names of all objects in your environment that contain a year (4 digits)
vehicle_names <- ls(pattern = "^vehicle\\d{4}")

# Create a named list of these data frames
vehicle_list <- mget(vehicle_names)

# Add year column to each vehicle data frame via function
vehicle_list_updated <- lapply(names(vehicle_list), function(df_name) {
  
  # Extract the year from the string (e.g., "2020")
  year_val <- str_extract(df_name, "\\d{4}")
  
  # Access the data frame
  df <- vehicle_list[[df_name]]
  
  # Add the column
  df$YEAR <- year_val
  
  return(df)
})

# 4. Rename the list elements to match original names
names(vehicle_list_updated) <- vehicle_names

# update R environment with 
list2env(vehicle_list_updated, envir = .GlobalEnv)
```

```{r}
#| label: vehicle-stacked
#| message: false
#| warning: false
#| echo: true  
#| cache: true

# define vehicle data frames in environment and convert all columns to characters
vehicle_list <- mget(ls(pattern = "^vehicle[0-9]")) |>  
  map(~ .x |>  mutate(across(everything(), as.character)))

# bind data frames
stacked_national_vehicle <- vehicle_list |>  
  bind_rows(.id = "source_year") |>  
  select(any_of(c(
    "ST_CASE", 
    "YEAR", 
    "STATE", 
    "COUNTY", 
    "VEH_NO", 
    "PER_NO", 
    "RUR_URB", 
    "RUR_URBNAME", 
    "PERSONS", 
    "FATALS", 
    "A_RU",
    "longitude", 
    "longitud", 
    "latitude", 
    "LATITUDE", 
    "LONGITUDE", 
    "LONGITUD",
### behavioral variables
    "DR_DRINK",
    "SPEEDREL",
    "PREV_SPD",
"DR_ZIP",
"SPEC_USE"
  )))

```

## Remove dataframes

I remove obsolete dataframes from my environment to clear up memory and to also remove unneeded dataframes for data exporting purposes.

```{r}
#| label: remove-dataframes-old
#| message: false
#| warning: false
#| echo: true  

# remove obsolete AUX data
rm(list = ls(pattern = "^acc20"))
rm(list = ls(pattern = "^per20"))
rm(list = ls(pattern = "^veh20"))

# remove obsolete National data
rm(list = ls(pattern = "accident20"))
rm(list = ls(pattern = "person20"))
rm(list = ls(pattern = "vehicle20"))
rm(list = ls(pattern = "_list"))
rm(list = c("person_names","vehicle_names"))


```

# Data Cleaning

| Stacked Dataset | Verify Correct Logic? | Notes |
|-------------------|-------------------|----------------------------------|
| stacked_national_person_v2 | ✅ | Not sure what logic to put for LAG_HRS due to the note |
| PER_AUX | ✅ |  |
| stacked_national_accident_v2 | ✅ |  |
| stacked_national_vehicle_v2 | ✅ |  |

## National person data cleaning

Okay this is now fixed. We need to keep the values as NA for now until we get to the export step. If we change the values to "" in this step then we run into bugs.

```{r}
#| label: national-person-data-cleaning
#| message: false
#| warning: false
#| echo: true  
#| cache: true

stacked_national_person_v2 <- stacked_national_person |>  
  mutate(RUR_URBNAME = if_else(RUR_URB >= 8 & RUR_URB <= 9, NA, RUR_URBNAME),
         RUR_URB = if_else(RUR_URB >= 8 & RUR_URB <= 9, NA, RUR_URB),
         REST_USE = if_else(REST_USE >= 96 & REST_USE <= 99, NA, REST_USE),
         REST_USENAME = if_else(REST_USE >= 96 & REST_USE <= 99, NA, REST_USENAME),
         REST_MIS = if_else(REST_MIS >= 8 & REST_MIS <= 9, NA, REST_MIS),
         REST_MISNAME = if_else(REST_MISNAME >= 8 & REST_MISNAME <= 9, NA, REST_MISNAME),
         HELM_USE = if_else(HELM_USE >= 96 & HELM_USE <= 99, NA, HELM_USE),
         HELM_USENAME = if_else(HELM_USENAME >= 96 & HELM_USENAME <= 99, NA, HELM_USENAME),
         BODY_TYP = if_else(BODY_TYP >= 98 & BODY_TYP <= 99, NA, BODY_TYP),
         BODY_TYPNAME = if_else(BODY_TYPNAME >= 98 & BODY_TYPNAME <= 99, NA, BODY_TYPNAME),
         INJ_SEV = if_else(INJ_SEV >= 8 & INJ_SEV <= 9, NA, INJ_SEV),
         INJ_SEVNAME = if_else(INJ_SEVNAME >= 8 & INJ_SEVNAME <= 9, NA, INJ_SEVNAME),
         SEX = if_else(SEX >= 8 & SEX <= 9, NA, SEX),
         AGE_2 = if_else(AGE >= 115, NA, as.character(AGE)),
         ALC_STATUS = if_else(ALC_STATUS >= 8 & ALC_STATUS <= 9, NA, ALC_STATUS),
         ALC_RES = if_else(ALC_RES > 94 & YEAR != 2015 | 
                           ALC_RES > 940 & YEAR == 2015, NA, ALC_RES),
         DRUGS = if_else(DRUGS >= 8 & DRUGS <= 9, NA, DRUGS),
         DSTATUS = if_else(DSTATUS >= 8 & DSTATUS <= 9, NA, DSTATUS),
         DOA = if_else(DOA == "9", NA, DOA),
         DEATH_MN = if_else(DEATH_MN > 59, NA, DEATH_MN),
         DEATH_DA = if_else(DEATH_DA > 31, NA, DEATH_DA),
         DEATH_YR = if_else(DEATH_YR > 2050, NA, DEATH_YR),
         LAG_HRS = if_else(LAG_HRS > 719, NA, LAG_HRS), ## note says: "/*Different before 2010*/". What should I put for logic?    
         LAG_MINS = if_else(LAG_MINS > 59, NA, LAG_MINS)
)

```

This step fixes the AGE character glitch in the previous step. I intentionally made this repetitive so I could check line by line if AGE is correct.

```{r}
#| label: stacked_national_person_v3
#| message: FALSE
#| warning: FALSE
#| echo: TRUE  

stacked_national_person_v3 <- stacked_national_person_v2 |> 
  mutate(AGE_3 = as.numeric(AGE)) |> 
  mutate(AGE_4 = if_else(AGE_3 >= 115, NA, AGE_3))

stacked_national_person_v3$AGE_4


test <- stacked_national_person_v3 |> 
  select(ST_CASE, AGE_4, AGE) |> 
  filter(is.na(AGE_4))
  

```

Okay, I confirm that this test is correct. The next thing to do is clean up stacked_national_person_v3, and also make sure the export replaces all NAs with blanks.

```{r}
#| label: stacked_national_person_v4
#| message: FALSE
#| warning: FALSE
#| echo: TRUE  

stacked_national_person_v4 <- stacked_national_person_v3 |> 
  select(-c(AGE, AGE_2, AGE_3))

stacked_national_person_v4$AGE_4
```

```{r}
#| label: remove-old-stacked-np
#| message: FALSE
#| warning: FALSE
#| echo: TRUE  

rm(stacked_national_person,
   stacked_national_person_v2,
   stacked_national_person_v3)
```


## AUX person files data cleaning

```{r}
#| label: AUX-person-data-cleaning
#| message: false
#| warning: false
#| echo: true  

stacked_AUX_per_v2 <- stacked_AUX_per |>  
  mutate(A_HRACE = if_else(A_HRACE == 9, "", as.character(A_HRACE)))

```

## National Accident files

I combined Trish data cleaning 2014 and 2015 as noted below.

```{r}
#| label: national-accident-data-cleaning
#| message: false
#| warning: false
#| echo: true  

stacked_national_accident_v2 <- stacked_national_accident |>  
  mutate(COUNTY = if_else(COUNTY > 996, "", COUNTY),
         LATITUDE = if_else(LATITUDE >= 77.0, "", LATITUDE),
         LONGITUD = if_else(LONGITUD > 700, "", LONGITUD),
         ROAD_FNC = if_else(
           ROAD_FNC == 9 & YEAR <= 2014 | 
           ROAD_FNC == 19 & YEAR <= 2014 | 
           ROAD_FNC == 99 & YEAR <= 2014, ## 2014 and before
           "", ROAD_FNC),
         FUNC_SYS = if_else(FUNC_SYS >= 96 & YEAR >= 2015, "", FUNC_SYS), ## 2015 and after
         DAY = if_else(DAY > 31, "", DAY),
         DAY_WEEK = if_else(DAY_WEEK > 7, "", DAY_WEEK),
         MONTH = if_else(MONTH >12, "", MONTH),
         HOUR = if_else(HOUR >24, "", HOUR),
         MINUTE = if_else(MINUTE > 59, "", MINUTE),
         NOT_HOUR = if_else(NOT_HOUR > 24, "", NOT_HOUR),
         NOT_MIN = if_else(NOT_MIN > 59, "", NOT_MIN),
         ARR_HOUR = if_else(ARR_HOUR > 24, "", ARR_HOUR),
         ARR_MIN = if_else(ARR_MIN > 59, "", ARR_MIN),
         HOSP_HR = if_else(HOSP_HR > 24, "", HOSP_HR),
         HOSP_MN = if_else(HOSP_MN > 59, "", HOSP_MN)
  )



```

## National Vehicle files

```{r}
#| label: label_name
#| message: false
#| warning: false
#| echo: true  

stacked_national_vehicle_v2 <- stacked_national_vehicle |>
  mutate(SPEEDREL = if_else( SPEEDREL == 9, "", SPEEDREL),
         PREV_SPD = if_else(PREV_SPD > 97, "", PREV_SPD),
         SPEC_USE = if_else(SPEC_USE>=98, "", SPEC_USE),
         DR_ZIP = if_else(DR_ZIP >= 99997, "", DR_ZIP)
  )

```

### Remove old stacked dataframes from environment

```{r}
#| label: remove-old-stacked-dataframes
#| message: false
#| warning: false
#| echo: true

rm(stacked_national_accident)
rm(stacked_national_person)
rm(stacked_national_vehicle)
rm(stacked_AUX_per)
```

# Merging Stacked Datasets

## Convert all variables to character

```{r}
#| label: convert-variables-character
#| message: false
#| warning: false
#| echo: true  

# 1. Identify all objects in the global environment that are data frames
df_names <- names(which(sapply(ls(), function(x) is.data.frame(get(x)))))

# 2. Loop through each data frame, convert columns, and save them back
for (name in df_names) {
  # Get the actual data frame object
  df <- get(name)
  
  # Convert all columns to character (using df[] preserves the data frame structure)
  df[] <- lapply(df, as.character)
  
  # Assign the modified data frame back to its original name in the global environment
  assign(name, df, envir = .GlobalEnv)
}
```

### remove values

```{r}
#| label: remove-values
#| message: false
#| warning: false
#| echo: true  

rm(df_names)
rm(name)
```

## Merge datasets

In this step, I merge the following datasets together using the stacked national person file as the base.

- AUX_person
- AUX_accident
- national_accident_v2

```{r}
#| label: merge-datasets
#| message: false
#| warning: false
#| echo: true  
#| cache: true


df <- stacked_national_person_v4  |>  
  left_join(stacked_national_accident_v2, by = c("ST_CASE", "YEAR")) |>  
  left_join(stacked_AUX_per_v2, by = c("ST_CASE","YEAR","VEH_NO","PER_NO")) |>  
  left_join(stacked_AUX_acc, by = c("ST_CASE","YEAR"))  |> 
  mutate(
         STATE = coalesce(STATE.x, STATE.y),
         COUNTY = coalesce(COUNTY.x, COUNTY.y),
         RUR_URB = coalesce(RUR_URB.x, RUR_URB.y),
         RUR_URBNAME = coalesce(RUR_URBNAME.x, RUR_URBNAME.y)) |> 
  select(-ends_with(".x"), -ends_with(".y"))


```


## Verification

I create a duplicate df without restrictions for verification with Trish.

```{r}
#| label: df-duplicate
#| message: false
#| warning: false
#| echo: true  

df_duplicate <- df
```

## Analytical Dataset restrictions

df is the completely merged dataset. I restrict the dataset to fatal cases in private vehicle uses and occupants inside cars.

```{r}
#| label: df-dataset-restrictions
#| message: false
#| warning: false
#| echo: true  

df <- df |>  
  filter(INJ_SEV == 4) |>   ## dead people only
  filter(YEAR >= 2010) |>   ## filter to years 2010+
## vehicle filters
  filter(BODY_TYP %in% c(1:8) |
          BODY_TYP == 17 | ## passenger cars
         BODY_TYP %in% c(14:16) | ## SUVs
           BODY_TYP == 19 |
         BODY_TYP == 20 | ## minivans
         BODY_TYP %in% c(30:39)) |>   ## pickups
## driver or occupant in car
  filter(A_PTYPE %in% c(1:2)) ##|>  
## private vehicle use [NOTE: CANNOT APPLY FILTER YET BECAUSE WE DID NOT MERGE IN NATIONAL VEHICLE FILE. WAITING FOR INSTRUCTIONS FROM TRISH]
 ## filter(SPEC_USE == 0)


```

# Troubleshoot

Need to fix issues before submitting to Trish.

## Binary Age Issue

It looks like the age issue is set to character and is binary. The final product is incorrect. Need to look back at the previous iterations and check if the issue persists there as well

```{r}
#| label: df-dup-age
#| message: FALSE
#| warning: FALSE
#| echo: TRUE  

# df_age_research <- df_duplicate |> 
#   select(ST_CASE, AGE) |> 
#   filter(AGE != "")

```

checking if the issue occurs at the row-binding step.

```{r}
#| label: age-stacked-national-person
#| message: FALSE
#| warning: FALSE
#| echo: TRUE  

# stacked_national_person_v2$AGE

```

Hmm it looks like this iteration still has issues. I wonder if the importing process makes the age column binary?

```{r}
#| label: age-import-stacked-np-fix
#| message: FALSE
#| warning: FALSE
#| echo: TRUE  

person2007_age <- fread("Raw Data/2007/FARS2007NationalCSV/PERSON.CSV") |> 
  select(ST_CASE, AGE) |> 
  filter(AGE != "")

person2007_age$AGE

```

```{r}
#| label: check-stacked-np
#| message: FALSE
#| warning: FALSE
#| echo: TRUE  

#stacked_national_person$AGE
```

Okay weird. It looks like the importing process is fine and stacked_national_person is fine.

It's most likely when the datasets are being merged. E.g when df is created

```{r}
#| label: age-fix
#| message: FALSE
#| warning: FALSE
#| echo: TRUE  

df$AGE_4


```

YES!!! The issue is when the data is being merged

# Data Export

I export all the datasets so I can upload them to BOX.

## df split

I split the dataset into three smaller subdatasets based on the following year ranges:

- 2010-2015
- 2016-2020
- 2021-2024

This is so we can verify the data in Excel.

```{r}
#| label: df-split
#| message: false
#| warning: false
#| echo: true  

df_2010_2015 <- df |>  
  filter(YEAR %in% c(2010:2015))

df_2016_2020 <- df |>  
  filter(YEAR %in% c(2016:2020))

df_2021_2024 <- df |>  
  filter(YEAR %in% c(2021:2024))

df_duplicate_2010_2015 <- df_duplicate |>  
  filter(YEAR %in% c(2010:2015))

df_duplicate_2016_2020 <- df_duplicate |>  
  filter(YEAR %in% c(2016:2020))

df_duplicate_2021_2024 <- df_duplicate |>  
  filter(YEAR %in% c(2021:2024))
```

## Remove Data Frames

```{r}
#| label: remove-data-frames
#| message: FALSE
#| warning: FALSE
#| echo: TRUE  

rm(list = ls(pattern = "stacked"),
   test)

```


## Data Export


```{r}
#| label: data_export
#| message: false
#| warning: false
#| echo: true
#| include: false ## remove after fixing

# I filter the environment for all data frame objects and create a list
all_data_frames <- mget(ls()) |>  
  keep(is.data.frame)


# I walk through the entire list and tell R to write .csv files into the "Export" folder.
 iwalk(all_data_frames, ~ write_csv(.x, paste0("Exports/Troubleshoot/", .y, ".csv"), na = ""))



```
